# Poverty Indices  {#poverty}

## At Risk of Poverty Ratio (svyarpr)


For additional usage examples of `svyarpr`, type `?convey::svyarpr` in the R console.


here are the references

[@osier2009] and [@deville1999]



## At Risk of Poverty Threshold (svyarpt)



For additional usage examples of `svylorenz`, type `?convey::svylorenz` in the R console.


here are the references

[@osier2009] and [@deville1999]



## Relative Median Income Ratio (svyrmir)


For additional usage examples of `svyrmir`, type `?convey::svyrmir` in the R console.


here are the references

[@osier2009] and [@deville1999]



## Relative Median Poverty Gap (svyrmpg)



For additional usage examples of `svyrmpg`, type `?convey::svyrmpg` in the R console.


here are the references

[@osier2009] and [@deville1999]


## Median Income Below the At Risk of Poverty Threshold  (svypoormed)


For additional usage examples of `svypoormed`, type `?convey::svypoormed` in the R console.


here are the references

[@osier2009] and [@deville1999]



## Foster-Greer-Thorbecke class (svyfgt)



[@foster1984] proposed a family of indicators to measure poverty.

The class of $FGT$ measures, can be defined as

\[
p=\frac{1}{N}\sum_{k\in U}h(y_{k},\theta ), 
\]

where

\[
h(y_{k},\theta )=\left[ \frac{(\theta -y_{k})}{\theta }\right] ^{\gamma
}\delta \left\{ y_{k}\leq \theta \right\} , 
\]

where: $\theta$ is the poverty threshold; $\delta$ the indicator function that assigns value 1 if the condition $\{y_{k}\leq \theta \}$ is satisfied and 0 otherwise, and $\gamma$ is a non-negative constant. 

When $\gamma =0$, $p$ can be interpreted as the poverty headcount ratio, and for $\gamma \geq 1$, the weight of the income shortfall of the poor to a power $\gamma$, (Foster and all, 1984).

The poverty measure FGT is implemented in the library convey by the function `svyfgt`.
The argument `thresh_type` of this function defines the type of poverty threshold adopted.
There are three possible choices:

1. `abs` -- fixed and given by the argument thresh_value
2. `relq` -- a proportion of a quantile fixed by the argument `proportion` and the quantile is defined by the argument `order`.
3. `relm` -- a proportion of the mean fixed the argument `proportion`

The quantile and the mean involved in the definition of the threshold are estimated for the whole population. When $\gamma=0$ and $\theta= .6*MED$ the measure is equal to the indicator `arpr` computed by
the function  `svyarpr`. 

Next, we give some examples of the function `svyfgt` to estimate the values of the FGT poverty index.

Consider first the poverty threshold fixed ($\gamma=0$) in the value $10000$. The headcount ratio (FGT0) is

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=0, abs_thresh=10000)
```

The poverty gap (FGT1) ($\gamma=1$) index for the poverty threshold fixed at the same value is

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=1, abs_thresh=10000)
```

To estimate the FGT0 with the poverty threshold fixed at $0.6* MED$ we fix the argument type_thresh="relq" and use the default values for  `percent`  and   `order`:

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=0, type_thresh= "relq")
```
that matches the estimate obtained by

```{r comment=NA}
svyarpr(~eqIncome, design=des_eusilc, .5, .6)
```
To estimate the poverty gap(FGT1) with the poverty threshold equal to $0.6*MEAN$
we use:

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=1, type_thresh= "relm")
```





For additional usage examples of `svyfgt`, type `?convey::svyfgt` in the R console.

here are the references

[@foster1984] and [@berger2003]


