# Poverty Indices  {#poverty}

## At Risk of Poverty Ratio (svyarpr)


For additional usage examples of `svyarpr`, type `?convey::svyarpr` in the R console.


here are the references

[@osier2009] and [@deville1999]



## At Risk of Poverty Threshold (svyarpt)



For additional usage examples of `svyarpt`, type `?convey::svyarpt` in the R console.


here are the references

[@osier2009] and [@deville1999]



## Relative Median Income Ratio (svyrmir)


For additional usage examples of `svyrmir`, type `?convey::svyrmir` in the R console.


here are the references

[@osier2009] and [@deville1999]



## Relative Median Poverty Gap (svyrmpg)



For additional usage examples of `svyrmpg`, type `?convey::svyrmpg` in the R console.


here are the references

[@osier2009] and [@deville1999]


## Median Income Below the At Risk of Poverty Threshold  (svypoormed)


For additional usage examples of `svypoormed`, type `?convey::svypoormed` in the R console.


here are the references

[@osier2009] and [@deville1999]



## Foster-Greer-Thorbecke class (svyfgt)



[@foster1984] proposed a family of indicators to measure poverty.  This class of $FGT$ measures, can be defined as

\[
p=\frac{1}{N}\sum_{k\in U}h(y_{k},\theta ), 
\]

where

\[
h(y_{k},\theta )=\left[ \frac{(\theta -y_{k})}{\theta }\right] ^{\gamma
}\delta \left\{ y_{k}\leq \theta \right\} , 
\]

where: $\theta$ is the poverty threshold; $\delta$ the indicator function that assigns value $1$ if the condition $\{y_{k}\leq \theta \}$ is satisfied and $0$ otherwise, and $\gamma$ is a non-negative constant. 

When $\gamma =0$, $p$ can be interpreted as the poverty headcount ratio, and for $\gamma \geq 1$, the weight of the income shortfall of the poor to a power $\gamma$, (Foster and all, 1984).

The poverty measure FGT is implemented in the library convey by the function `svyfgt`.
The argument `thresh_type` of this function defines the type of poverty threshold adopted.
There are three possible choices:

1. `abs` -- fixed and given by the argument thresh_value
2. `relq` -- a proportion of a quantile fixed by the argument `proportion` and the quantile is defined by the argument `order`.
3. `relm` -- a proportion of the mean fixed the argument `proportion`

The quantile and the mean involved in the definition of the threshold are estimated for the whole population. When $\gamma=0$ and $\theta= .6*MED$ the measure is equal to the indicator `arpr` computed by
the function  `svyarpr`. 

Next, we give some examples of the function `svyfgt` to estimate the values of the FGT poverty index.

Consider first the poverty threshold fixed ($\gamma=0$) in the value $10000$. The headcount ratio (FGT0) is

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=0, abs_thresh=10000)
```

The poverty gap (FGT1) ($\gamma=1$) index for the poverty threshold fixed at the same value is

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=1, abs_thresh=10000)
```

To estimate the FGT0 with the poverty threshold fixed at $0.6* MED$ we fix the argument type_thresh="relq" and use the default values for  `percent`  and   `order`:

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=0, type_thresh= "relq")
```
that matches the estimate obtained by

```{r comment=NA}
svyarpr(~eqIncome, design=des_eusilc, .5, .6)
```
To estimate the poverty gap(FGT1) with the poverty threshold equal to $0.6*MEAN$
we use:

```{r comment=NA}
svyfgt(~eqIncome, des_eusilc, g=1, type_thresh= "relm")
```


---

**A replication example**

In July 2006, [@jenkins2006] presented at the North American Stata Users' Group Meetings on the stata Atkinson Index command.  The example below reproduces those statistics.

Load and prepare the same data set:
```{r}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the foreign library
library(foreign)

# create a temporary file on the local disk
tf <- tempfile()

# store the location of the presentation file
presentation_zip <- "http://repec.org/nasug2006/nasug2006_jenkins.zip"

# download jenkins' presentation to the temporary file
download.file( presentation_zip , tf , mode = 'wb' )

# unzip the contents of the archive
presentation_files <- unzip( tf , exdir = tempdir() )

# load the institute for fiscal studies' 1981, 1985, and 1991 data.frame objects
x81 <- read.dta( grep( "ifs81" , presentation_files , value = TRUE ) )
x85 <- read.dta( grep( "ifs85" , presentation_files , value = TRUE ) )
x91 <- read.dta( grep( "ifs91" , presentation_files , value = TRUE ) )

# NOTE: we recommend using ?convey::svyarpt rather than this unweighted calculation #
# Why not to use svyquantile: can both the correct estimate and the correct se#

# calculate 60% of the unweighted median income in 1981
unwtd_arpt81 <- quantile( x81$eybhc0 , 0.5 ) * .6

# calculate 60% of the unweighted median income in 1985
unwtd_arpt85 <- quantile( x85$eybhc0 , 0.5 ) * .6

# calculate 60% of the unweighted median income in 1991
unwtd_arpt91 <- quantile( x91$eybhc0 , 0.5 ) * .6

# stack each of these three years of data into a single data.frame
x <- rbind( x81 , x85 , x91 )
```

Replicate the author's survey design statement from stata code..
```{code}
. ge poor = (year==1981)*(x < $z_81) + (year==1985)*(x < $z_85) +  (year==1991)*(x < $z_91)
. * account for clustering within HHs 
. svyset hrn [pweight = wgt]
```

.. into R code:


```{r}
# initiate a linearized survey design object
y <- svydesign( ~ hrn , data = x , weights = ~ wgt )

# immediately run the `convey_prep` function on the survey design
z <- convey_prep( y )
```

Replicate the author's headcount ratio results with stata..
```{code}
. svy: mean poor if year == 1981
(running mean on estimation sample)

Survey: Mean estimation

Number of strata =       1          Number of obs    =    9772
Number of PSUs   =    7476          Population size  = 5.5e+07
                                    Design df        =    7475

--------------------------------------------------------------
             |             Linearized
             |       Mean   Std. Err.     [95% Conf. Interval]
-------------+------------------------------------------------
        poor |   .1410125   .0044859       .132219     .149806
--------------------------------------------------------------

. svy: mean poor if year == 1985
(running mean on estimation sample)

Survey: Mean estimation

Number of strata =       1          Number of obs    =    8991
Number of PSUs   =    6972          Population size  = 5.5e+07
                                    Design df        =    6971

--------------------------------------------------------------
             |             Linearized
             |       Mean   Std. Err.     [95% Conf. Interval]
-------------+------------------------------------------------
        poor |    .137645   .0046531      .1285235    .1467665
--------------------------------------------------------------

. svy: mean poor if year == 1991
(running mean on estimation sample)

Survey: Mean estimation

Number of strata =       1          Number of obs    =    6468
Number of PSUs   =    5254          Population size  = 5.6e+07
                                    Design df        =    5253

--------------------------------------------------------------
             |             Linearized
             |       Mean   Std. Err.     [95% Conf. Interval]
-------------+------------------------------------------------
        poor |   .2021312   .0062077      .1899615    .2143009
--------------------------------------------------------------
```

..using R code:

```{r}
headcount_81 <- 
	svyfgt( 
		~ eybhc0 , 
		subset( z , year == 1981 ) , 
		g = 0 , 
		abs_thresh = unwtd_arpt81
	)

headcount_81

confint( headcount_81 , df = degf( subset( z , year == 1981 ) ) )

headcount_85 <- 
	svyfgt( 
		~ eybhc0 , 
		subset( z , year == 1985 ) , 
		g = 0 , 
		abs_thresh = unwtd_arpt85 
	)
	
headcount_85

confint( headcount_85 , df = degf( subset( z , year == 1985 ) ) )

headcount_91 <- 
	svyfgt( 
		~ eybhc0 , 
		subset( z , year == 1991 ) , 
		g = 0 , 
		abs_thresh = unwtd_arpt91 
	)

headcount_91
	
confint( headcount_91 , df = degf( subset( z , year == 1991 ) ) )
```



Confirm this replication applies for the normalized poverty gap as well, comparing stata code..
```{code}
. ge ngap = poor*($z_81- x)/$z_81 if year == 1981

. svy: mean ngap if year == 1981
(running mean on estimation sample)

Survey: Mean estimation

Number of strata =       1          Number of obs    =    9772
Number of PSUs   =    7476          Population size  = 5.5e+07
                                    Design df        =    7475

--------------------------------------------------------------
             |             Linearized
             |       Mean   Std. Err.     [95% Conf. Interval]
-------------+------------------------------------------------
        ngap |   .0271577   .0013502      .0245109    .0298044
--------------------------------------------------------------
```

..to R code:

```{r}
norm_pov_81 <- 
	svyfgt( 
		~ eybhc0 , 
		subset( z , year == 1981 ) , 
		g = 1 , 
		abs_thresh = unwtd_arpt81
	)
	
norm_pov_81

confint( norm_pov_81 , df = degf( subset( z , year == 1981 ) ) )
```





For additional usage examples of `svyfgt`, type `?convey::svyfgt` in the R console.

here are the references

[@foster1984] and [@berger2003]


