# Inequality Measurement {#inequality}

[add brief explanation on concentration, Lorenz curve and the beggining of inequality measurement]

## Entropy-based Measures

Entropy is a concept derived from information theory, meaning the expected amount of information given the occurrence of an event. Following [@shannon1948], given an event $y$ with probability density function $f(\cdot)$, the information content given the occurrence of $y$ can be defined as $g(f(y)) \colon= - \log f(y)$. Therefore, the expected information or, put simply, the *entropy* is

\[
H(f) \colon = -E \big[ \log f(y) \big] = - \int_{-\infty}^{\infty} f(y) \log f(y) dy
\]

Assuming a discrete distribution, with $p_k$ as the probability of occurring event $k \in K$, the entropy formula takes the form:

\[
H = - \sum_{k \in K} p_k \log p_k \text{.}
\]

The main idea behind it is that the expected amount of information of an event is inversely proportional to the probability of its occurrence. In other words, the information derived from the observation of a rare event is higher than of the information of more probable events.

Using the intuition presented in [@cowell2009], substituting the density function by the income share of an individual $s(q) = {F}^{-1}(q) / \int_{0}^{1} F^{-1}(t)dt = y/\mu$, the entropy function becomes the Theil inequality index

\[
I_{Theil} = \int_{0}^{\infty} \frac{y}{\mu} \log \bigg( \frac{y}{\mu} \bigg) dF(y) = -H(s)
\]

Therefore, the entropy-based inequality measure increases as a person's income $y$ deviates from the mean $\mu$. This is the basic idea behind entropy-based inequality measures.

## Lorenz Curve (svylorenz)

Though not an inequality measure in itself, the Lorenz curve is a classic instrument of distribution analysis. Basically, it is a function that associates a cumulative share of the population to the share of the total income it owns. In mathematical terms, 

\[
L(p) = \frac{\int_{-\infty}^{Q_p}yf(y)dy}{\int_{-\infty}^{+\infty}yf(y)dy}
\]

where $Q_p$ is the quantile $p$ of the population.

The two extreme distributive cases are 

- Perfect equality:
    - Every individual has the same income;
    - Every share of the population has the same share of the income;
    - Therefore, the reference curve is \[L(p) = p \text{ } \forall p \in [0,1] \text{.}\]
- Perfect inequality:
    - One individual concentrates all of society's income, while the other individuals have zero income;
    - Therefore, the reference curve is 
    
\[
L(p)=
\begin{cases}
0, &\forall p < 1 \\
1, &\text{if } p = 1 \text{.}
\end{cases}
\]
    
In order to evaluate the degree of inequality in a society, the analyst looks at the distance between the real curve and those two reference curves.

The estimator of this function was derived by [@kovacevic1997]:

\[
L(p) = \frac{ \sum_{i \in S} w_i \cdot y_i \cdot \delta \{ y_i \le \widehat{Q}_p \}}{\widehat{Y}}, \text{ } 0 \le p \le 1.
\]

Yet, this formula is used to calculate specific points of the curve and their respective SEs. The formula to plot an approximation of the continuous empirical curve comes from [@lerman1989].

---

**A replication example**

In October 2016, [@jann2016] released a pre-publication working paper to estimate lorenz and concentration curves using stata.  The example below reproduces the statistics presented in his section 4.1.

```{r}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the stata-style webuse library
library(webuse)

# load the NLSW 1988 data
webuse("nlsw88")

# coerce that `tbl_df` to a standard R `data.frame`
nlsw88 <- data.frame( nlsw88 )

# initiate a linearized survey design object
des_nlsw88 <- svydesign( ids = ~1 , data = nlsw88 )

# immediately run the `convey_prep` function on the survey design
des_nlsw88 <- convey_prep(des_nlsw88)

# estimates lorenz curve
result.lin <- svylorenz( ~wage, des_nlsw88, quantiles = seq( 0, 1, .05 ), na.rm = T )


# note: most survey commands in R use Inf degrees of freedom by default
# stata generally uses the degrees of freedom of the survey design.
# therefore, while this extended syntax serves to prove a precise replication of stata
# it is generally not necessary.
section_four_one <-
	data.frame( 
		estimate = coef( result.lin ) , 
		standard_error = SE( result.lin ) , 
		ci_lower_bound = 
		    coef( result.lin ) + 
		    SE( result.lin ) * 
		    qt( 0.025 , degf( subset( des_nlsw88 , !is.na( wage ) ) ) ) ,
		ci_upper_bound = 
		    coef( result.lin ) + 
		    SE( result.lin ) * 
		    qt( 0.975 , degf( subset( des_nlsw88 , !is.na( wage ) ) ) )
	)
	

knitr::kable(
  section_four_one , caption = 'Here is a nice table!',
  booktabs = TRUE
)
```


For additional usage examples of `svylorenz`, type `?convey::svylorenz` in the R console.


## Gini index (svygini)

The Gini index is an attempt to express the inequality presented in the Lorenz curve as a single number. In essence, it is twice the area between the equality curve and the real Lorenz curve. Put simply:

\[
\begin{aligned}
G &= 2 \bigg( \int_{0}^{1} pdp - \int_{0}^{1} L(p)dp \bigg) \\
\therefore G &= 1 - 2 \int_{0}^{1} L(p)dp
\end{aligned}
\]

where $G=0$ in case of perfect equality and $G = 1$ in the case of perfect inequality.

The estimator proposed by [@osier2009] is defined as:

\[
\widehat{G} = \frac{ 2 \sum_{i \in S} w_i r_i y_i - \sum_{i \in S} w_i y_i }{ \hat{Y} }
\]

The linearized formula of $\widehat{G}$ is used to calculate the SE.


For additional usage examples of `svygini`, type `?convey::svygini` in the R console.


## Amato index (svyamato)

The Amato index is also based on the Lorenz curve, but instead of focusing on the area of the curve, it focuses on its length. [@arnold2012] proposes a formula not directly based in the Lorenz curve, which [@barabesi2016] uses to present the following estimator:

\[
\widehat{A} = \sum_{i \in S} w_i \bigg[ \frac{1}{\widehat{N}^2} + \frac{y_i^2}{\widehat{Y}^2} \bigg]^{\frac{1}{2}} \text{,}
\]

which also generates the linearized formula for SE estimation.

The minimum value $A$ assumes is $\sqrt{2}$ and the maximum is $2$. In order to get a measure in the interval $[0,1]$, the standardized Amato index $\widetilde{A}$ can be defined as:

\[
\widetilde{A} = \frac{ A - \sqrt{2} }{2 - \sqrt{2} } \text{ .}
\]


For additional usage examples of `svyamato`, type `?convey::svyamato` in the R console.


## Zenga Index and Curve (svyzenga, svyzengacurve)

The Zenga index and its curve were proposed in [@zenga2007]. As [@polisicchio2011] noticed, this curve derives directly from the Lorenz curve, and can be defined as:

\[
Z(p) = 1 - \frac{L(p)}{p} \cdot \frac{1 - p}{1 - L(p)}.
\]

In the `convey` library, an experimental estimator based on the Lorenz curve is used:

\[
\widehat{Z(p)} = \frac{ p \widehat{Y} - \widehat{\widetilde{Y}}(p) }{p \big[ \widehat{Y} - \widehat{\widetilde{Y}}(p) \big] }.
\]

In turn, the Zenga index derives from this curve and is defined as:

\[
Z = \int_0^1 Z(p)dp.
\]

However, its estimators were proposed by [@langel2012] and [@barabesi2016]. In this library, the latter is used and is defined as:

\[
\widehat{Z} = 1 - \sum_{i \in S} w_i \bigg[ \frac{ ( \widehat{N} - \widehat{H}_{y_i} ) ( \widehat{Y} -\widehat{K}_{y_i} ) }
{ \widehat{N} \cdot \widehat{H}_{y_i} \cdot \widehat{K}_{y_i} } \bigg]
\]

where $\widehat{N}$ is the population total, $\widehat{Y}$ is the total income, $\widehat{H}_{y_i}$ is the sum of incomes below or equal to $y_i$ and $\widehat{N}_{y_i}$ is the sum of incomes greater or equal to $y_i$.



For additional usage examples of `svyzenga` or `svyzengacurve`, type `?convey::svyzenga` or `?convey::svyzengacurve` in the R console.



## Generalized Entropy and Decomposition (svygei, svygeidec)

Using a generalization of the information function, now defined as $g(f) = \frac{1}{\alpha-1} [ 1 - f^{\alpha - 1} ]$, the $\alpha$-class entropy is 
\[
H_\alpha(f) = \frac{1}{\alpha - 1} \bigg[ 1 - \int_{-\infty}^{\infty} f(y)^{ \alpha - 1} f(y) dy \bigg] \text{.}
\]

This relates to a class of inequality measures, the Generalized entropy indices, defined as:

\[
GE_\alpha = \frac{1}{\alpha^2 - \alpha} \int_{0}^\infty \bigg[ \bigg( \frac{y}{\mu} \bigg)^\alpha - 1 \bigg]dF(x) = - \frac{-H_\alpha(s) }{ \alpha } \text{.}
\]

The parameter $\alpha$ also has an economic interpretation: as $\alpha$ increases, the influence of top incomes upon the index increases. In some cases, this measure takes special forms, such as mean log deviation and the aforementioned Theil index.

In order to estimate it, [@biewen2003] proposed the following:

\[
GE_\alpha =
\begin{cases}
( \alpha^2 - \alpha)^{-1} \big[ U_0^{\alpha - 1} U_1^{-\alpha} U_\alpha -1 \big], & \text{if } \alpha \in \mathbb{R} \setminus \{0,1\} \\
- T_0 U_0^{-1} + \log ( U_1 / U_0 ), &\text{if } \alpha \rightarrow 0 \\
T_1 U_1^{-1} - \log ( U_1 / U_0 ), & \text{if } \alpha \rightarrow 1
\end{cases}
\]

where $U_\gamma = \sum_{i \in S} w_i \cdot y_i^\gamma$ and $T_\gamma = \sum_{i \in S} w_i \cdot y_i^\gamma \cdot \log y_i$. since those are all functions of totals, the linearization of the indices are easily achieved using the theorems described in [@deville1999].

This class also has several desirable properties, such as additive decomposition. The additive decomposition allows to compare the effects of inequality within and between population groups on the population inequality. Put simply, an additive decomposable index allows for:

\[
I_{Total} = I_{Between} + I_{Within} \text{.}
\]

---

**A replication example**

In July 2006, [@jenkins2006] presented at the North American Stata Users' Group Meetings on the stata Generalized Entropy Index command.  The example below reproduces those statistics.

Load and prepare the same data set:
```{r}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the foreign library
library(foreign)

# create a temporary file on the local disk
tf <- tempfile()

# store the location of the presentation file
presentation_zip <- "http://repec.org/nasug2006/nasug2006_jenkins.zip"

# download jenkins' presentation to the temporary file
download.file( presentation_zip , tf , mode = 'wb' )

# unzip the contents of the archive
presentation_files <- unzip( tf , exdir = tempdir() )

# load the institute for fiscal studies' 1981, 1985, and 1991 data.frame objects
x81 <- read.dta( grep( "ifs81" , presentation_files , value = TRUE ) )
x85 <- read.dta( grep( "ifs85" , presentation_files , value = TRUE ) )
x91 <- read.dta( grep( "ifs91" , presentation_files , value = TRUE ) )

# stack each of these three years of data into a single data.frame
x <- rbind( x81 , x85 , x91 )
```

Replicate the author's survey design statement from stata code..
```{code}
. * account for clustering within HHs 
. version 8: svyset [pweight = wgt], psu(hrn)
pweight is wgt
psu is hrn
construct an
```

.. into R code:


```{r}
# initiate a linearized survey design object
y <- svydesign( ~ hrn , data = x , weights = ~ wgt )

# immediately run the `convey_prep` function on the survey design
z <- convey_prep( y )
```

Replicate the author's subset statement and each of his svygei results..
```{code}
. svygei x if year == 1981
 
Warning: x has 20 values = 0. Not used in calculations

Complex survey estimates of Generalized Entropy inequality indices
 
pweight: wgt                                   Number of obs    = 9752
Strata: <one>                                  Number of strata = 1
PSU: hrn                                       Number of PSUs   = 7459
											   Population size  = 54766261
---------------------------------------------------------------------------
Index    |  Estimate   Std. Err.      z      P>|z|     [95% Conf. Interval]
---------+-----------------------------------------------------------------
GE(-1)   |  .1902062   .02474921     7.69    0.000      .1416987   .2387138
MLD      |  .1142851   .00275138    41.54    0.000      .1088925   .1196777
Theil    |  .1116923   .00226489    49.31    0.000      .1072532   .1161314
GE(2)    |   .128793   .00330774    38.94    0.000      .1223099    .135276
GE(3)    |  .1739994   .00662015    26.28    0.000      .1610242   .1869747
---------------------------------------------------------------------------
```

..using R code:

```{r}
z81 <- subset( z , year == 1981 )

svygei( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = -1 )
svygei( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = 0 )
svygei( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) )
svygei( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = 2 )
svygei( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = 3 )
```	



Confirm this replication applies for subsetted objects as well.  Compare stata output..

```{code}
. svygei x if year == 1985 & x >= 1

Complex survey estimates of Generalized Entropy inequality indices
 
pweight: wgt                                   Number of obs    = 8969
Strata: <one>                                  Number of strata = 1
PSU: hrn                                       Number of PSUs   = 6950
											   Population size  = 55042871
---------------------------------------------------------------------------
Index    |  Estimate   Std. Err.      z      P>|z|     [95% Conf. Interval]
---------+-----------------------------------------------------------------
GE(-1)   |  .1602358   .00936931    17.10    0.000      .1418723   .1785993
MLD      |   .127616   .00332187    38.42    0.000      .1211052   .1341267
Theil    |  .1337177   .00406302    32.91    0.000      .1257543    .141681
GE(2)    |  .1676393   .00730057    22.96    0.000      .1533304   .1819481
GE(3)    |  .2609507   .01850689    14.10    0.000      .2246779   .2972235
---------------------------------------------------------------------------
```

..to R code:

```{r}
z85 <- subset( z , year == 1985 )

svygei( ~ eybhc0 , subset( z85 , eybhc0 > 1 ) , epsilon = -1 )
svygei( ~ eybhc0 , subset( z85 , eybhc0 > 1 ) , epsilon = 0 )
svygei( ~ eybhc0 , subset( z85 , eybhc0 > 1 ) )
svygei( ~ eybhc0 , subset( z85 , eybhc0 > 1 ) , epsilon = 2 )
svygei( ~ eybhc0 , subset( z85 , eybhc0 > 1 ) , epsilon = 3 )
```



For additional usage examples of `svygei` or `svygeidec`, type `?convey::svygei` or `?convey::svygeidec` in the R console.



## Rényi Divergence (svyrenyi)

Another measure used in areas like ecology, statistics and information theory is Rényi divergence measure. Using the formula defined in [@langel2012], the estimator can be defined as:

\[
\widehat{R}_\alpha =
\begin{cases}
\frac{1}{\alpha - 1} \log \bigg[ \widehat{N}^{\alpha - 1} \sum_{i \in S} w_i \cdot \bigg( \frac{y_k}{ \widehat{Y} } \bigg) \bigg], &\text{if } \alpha \neq 1, \\
\sum_{i \in S} \frac{w_i y_i}{ \widehat{Y}} \log \frac{\widehat{N} y_i}{\widehat{Y}}, &\text{if } \alpha = 1,
\end{cases}
\]

where $\alpha$ is a parameter with a similar economic interpretation to that of the $GE_\alpha$ index.

## J-Divergence and Decomposition (svyjdiv, svyjdivdec)

Proposed by [@rohde2016], the J-divergence measure can be seen as the sum of $GE_0$ and $GE_1$, satisfying axioms that, individually, those two indices do not. Using $U_\gamma$ and $T_\gamma$ functions defined in \ref{subsection.3.3.1}, the estimator can be defined as:

\[
\begin{aligned}
\widehat{J} &= \frac{1}{\widehat{N}} \sum_{i \in S} w_i \bigg( \frac{ y_i - \widehat{\mu} }{ \widehat{\mu} } \bigg) \log \bigg( \frac{y_i}{\widehat{\mu}} \bigg) \\
\therefore \widehat{J} &= \frac{\widehat{T}_1}{\widehat{U}_1} - \frac{ \widehat{T}_0 }{ \widehat{U}_0 }
\end{aligned}
\]

Since it is a sum of two additive decomposable measures, $J$ itself is decomposable.





For additional usage examples of `svyrenyi`, type `?convey::svyrenyi` in the R console.


## Atkinson index (svyatk)

Although the original formula was proposed in [@atkinson1970], the estimator used here comes from [@biewen2003]:

\[
\widehat{A}_\epsilon =
\begin{cases}
 1 - \widehat{U}_0^{ - \epsilon/(1 - \epsilon) } \widehat{U}_1^{ -1 } \widehat{U}_{1 - \epsilon}^{ 1/(1 - \epsilon) } , &\text{if } \epsilon \in \mathbb{R}_+ \setminus\{ 1 \} \\
1 - \widehat{U}_0 \widehat{U}_0^{-1} exp( \widehat{T}_0 \widehat{U}_0^{-1} ), &\text{if } \epsilon \rightarrow1
\end{cases}
\]

The $\epsilon$ is an inequality aversion parameter: as it approaches infinity, more weight is given to incomes in bottom of the distribution.

---

**A replication example**

In July 2006, [@jenkins2006] presented at the North American Stata Users' Group Meetings on the stata Atkinson Index command.  The example below reproduces those statistics.

Load and prepare the same data set:
```{r}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the foreign library
library(foreign)

# create a temporary file on the local disk
tf <- tempfile()

# store the location of the presentation file
presentation_zip <- "http://repec.org/nasug2006/nasug2006_jenkins.zip"

# download jenkins' presentation to the temporary file
download.file( presentation_zip , tf , mode = 'wb' )

# unzip the contents of the archive
presentation_files <- unzip( tf , exdir = tempdir() )

# load the institute for fiscal studies' 1981, 1985, and 1991 data.frame objects
x81 <- read.dta( grep( "ifs81" , presentation_files , value = TRUE ) )
x85 <- read.dta( grep( "ifs85" , presentation_files , value = TRUE ) )
x91 <- read.dta( grep( "ifs91" , presentation_files , value = TRUE ) )

# stack each of these three years of data into a single data.frame
x <- rbind( x81 , x85 , x91 )
```

Replicate the author's survey design statement from stata code..
```{code}
. * account for clustering within HHs 
. version 8: svyset [pweight = wgt], psu(hrn)
pweight is wgt
psu is hrn
construct an
```

.. into R code:


```{r}
# initiate a linearized survey design object
y <- svydesign( ~ hrn , data = x , weights = ~ wgt )

# immediately run the `convey_prep` function on the survey design
z <- convey_prep( y )
```

Replicate the author's subset statement and each of his svyatk results with stata..
```{code}
. svyatk x if year == 1981
 
Warning: x has 20 values = 0. Not used in calculations

Complex survey estimates of Atkinson inequality indices
 
pweight: wgt                                   Number of obs    = 9752
Strata: <one>                                  Number of strata = 1
PSU: hrn                                       Number of PSUs   = 7459
                                               Population size  = 54766261
---------------------------------------------------------------------------
Index    |  Estimate   Std. Err.      z      P>|z|     [95% Conf. Interval]
---------+-----------------------------------------------------------------
A(0.5)   |  .0543239   .00107583    50.49    0.000      .0522153   .0564324
A(1)     |  .1079964   .00245424    44.00    0.000      .1031862   .1128066
A(1.5)   |  .1701794   .0066943    25.42    0.000       .1570588      .1833
A(2)     |  .2755788   .02597608    10.61    0.000      .2246666    .326491
A(2.5)   |  .4992701   .06754311     7.39    0.000       .366888   .6316522
---------------------------------------------------------------------------
```

..using R code:

```{r}
z81 <- subset( z , year == 1981 )

svyatk( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = 0.5 )
svyatk( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) )
svyatk( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = 1.5 )
svyatk( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = 2 )
svyatk( ~ eybhc0 , subset( z81 , eybhc0 > 0 ) , epsilon = 2.5 )
```



Confirm this replication applies for subsetted objects as well, comparing stata code..
```{code}
. svyatk x if year == 1981 & x >= 1

Complex survey estimates of Atkinson inequality indices
 
pweight: wgt                                   Number of obs    = 9748
Strata: <one>                                  Number of strata = 1
PSU: hrn                                       Number of PSUs   = 7457
                                               Population size  = 54744234
---------------------------------------------------------------------------
Index    |  Estimate   Std. Err.      z      P>|z|     [95% Conf. Interval]
---------+-----------------------------------------------------------------
A(0.5)   |  .0540059   .00105011    51.43    0.000      .0519477   .0560641
A(1)     |  .1066082   .00223318    47.74    0.000      .1022313   .1109852
A(1.5)   |  .1638299   .00483069    33.91    0.000       .154362   .1732979
A(2)     |  .2443206   .01425258    17.14    0.000      .2163861   .2722552
A(2.5)   |   .394787   .04155221     9.50    0.000      .3133461   .4762278
---------------------------------------------------------------------------
```

..to R code:

```{r}
z81_two <- subset( z , year == 1981 & eybhc0 > 1 )

svyatk( ~ eybhc0 , z81_two , epsilon = 0.5 )
svyatk( ~ eybhc0 , z81_two )
svyatk( ~ eybhc0 , z81_two , epsilon = 1.5 )
svyatk( ~ eybhc0 , z81_two , epsilon = 2 )
svyatk( ~ eybhc0 , z81_two , epsilon = 2.5 )
```




For additional usage examples of `svyatk`, type `?convey::svyatk` in the R console.
