---
output:
  pdf_document: default
  html_document: default
---
# Multidimensional Indices {#multidimensional}

## Alkire-Foster Class and Decomposition (svyafc, svyafcdec)



#### replication example

In November 2015, Christopher Jindra presented at the Oxford Poverty and Human Development Initiative on the Alkire-Foster multidimensional poverty measure.  His presentation can be viewed [here](http://www.ophi.org.uk/wp-content/uploads/Jindra_151109_OPHISeminar.pdf).  The example below reproduces those statistics.

Load and prepare the same data set:
```{r message=FALSE, warning=FALSE}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the stata-style webuse library
library(webuse)

# load the same microdata set used by Jindra in his presentation
webuse("nlsw88")

# coerce that `tbl_df` to a standard R `data.frame`
nlsw88 <- data.frame( nlsw88 )

# create a `collgrad` column
nlsw88$collgrad <-
    factor( 
        as.numeric( nlsw88$collgrad ) , 
        label = c( 'not college grad' , 'college grad' ) , 
        ordered = TRUE 
      )

# initiate a linearized survey design object
des_nlsw88 <- svydesign( ids = ~1 , data = nlsw88 )

# immediately run the `convey_prep` function on the survey design
des_nlsw88 <- convey_prep(des_nlsw88)
```


Replicate PDF page 9
```{r message=FALSE, warning=FALSE}
page_nine <-
  svyafc(
    ~ wage + collgrad + hours , 
    design = des_nlsw88 , 
    cutoffs = list( 4, 'college grad' , 26 ) , 
    k = 1/3 , g = 0 , 
    na.rm = TRUE
  )

# MO and seMO
print( page_nine )

# H seH and A seA
print( attr( page_nine , "extra" ) )
```


Replicate PDF page 10
```{r message=FALSE, warning=FALSE}
page_ten <- NULL

# loop through every poverty cutoff `k`
for( ks in seq( 0.1 , 1 , .1 ) ){
	
	this_ks <-
	    svyafc(
	        ~ wage + collgrad + hours , 
	        design = des_nlsw88 , 
	        cutoffs = list( 4 , 'college grad' , 26 ) , 
	        k = ks , 
	        g = 0 , 
	        na.rm = TRUE 
	       )
	
	page_ten <-
	    rbind(
	        page_ten ,
	        data.frame( 
	            k = ks , 
	            MO = coef( this_ks ) ,
	            seMO = SE( this_ks ) ,
	            H = attr( this_ks , "extra" )[ 1 , 1 ] ,
	            seH = attr( this_ks , "extra" )[ 1 , 2 ] ,
	            A = attr( this_ks , "extra" )[ 2 , 1 ] ,
	            seA = attr( this_ks , "extra" )[ 2 , 2 ]
          )
	    )
	
}

knitr::kable(
  page_ten , caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

still need to replicate PDF page 13

https://github.com/DjalmaPessoa/convey/issues/168


then keep going replicating this

https://github.com/DjalmaPessoa/convey/issues/154




[@alkire2011] and [@alkire2015] and [@pacifico2016]



## Bourguignon (1999) inequality class (svybmi)


[@bourguignon1999] and [@lugo2007]
