---
output:
  pdf_document: default
  html_document: default
---
# Multidimensional Indices {#multidimensional}

Inequality and poverty can be seen as multidimensional concepts, combining several livelihood characteristics. Usual approaches take into account income, housing, sanitation, etc. 

In order to transform these different measures from into meaningful numbers, economic theory builds on the idea of utility functions. Utility is a measure of well-being, assigning a "well-being score" to a vector of characteristics. Depending on the utility function, the analyst may allow for substitutions among characteristics: for instance, someone with a slightly lower income, but with access to sanitation, can have a higher wellbeing than someone with a higher income, but without access to sanitation. This depends on the set of weights given to the set of attributes.

Most measures below follow from this kind of two-step procedure: (1) estimating individual scores from an individual's set of characteristics; then (2) aggregating those individual scores into a single measure for the population.

The following section will present a measure of multidimensional poverty and a measure of multidimensional inequality, describing the main aspects of the theory and estimation procedures of each.

## Alkire-Foster Class and Decomposition (svyafc, svyafcdec)



---

**A replication example**

In November 2015, Christopher Jindra presented at the Oxford Poverty and Human Development Initiative on the Alkire-Foster multidimensional poverty measure.  His presentation can be viewed [here](http://www.ophi.org.uk/wp-content/uploads/Jindra_151109_OPHISeminar.pdf).  The example below reproduces those statistics.

Load and prepare the same data set:
```{r message=FALSE, warning=FALSE}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the stata-style webuse library
library(webuse)

# load the same microdata set used by Jindra in his presentation
webuse("nlsw88")

# coerce that `tbl_df` to a standard R `data.frame`
nlsw88 <- data.frame( nlsw88 )

# create a `collgrad` column
nlsw88$collgrad <-
    factor( 
        as.numeric( nlsw88$collgrad ) , 
        label = c( 'not college grad' , 'college grad' ) , 
        ordered = TRUE 
      )

# initiate a linearized survey design object
des_nlsw88 <- svydesign( ids = ~1 , data = nlsw88 )

# immediately run the `convey_prep` function on the survey design
des_nlsw88 <- convey_prep(des_nlsw88)
```


Replicate PDF page 9
```{r message=FALSE, warning=FALSE}
page_nine <-
  svyafc(
    ~ wage + collgrad + hours , 
    design = des_nlsw88 , 
    cutoffs = list( 4, 'college grad' , 26 ) , 
    k = 1/3 , g = 0 , 
    na.rm = TRUE
  )

# MO and seMO
print( page_nine )

# H seH and A seA
print( attr( page_nine , "extra" ) )
```


Replicate PDF page 10
```{r message=FALSE, warning=FALSE}
page_ten <- NULL

# loop through every poverty cutoff `k`
for( ks in seq( 0.1 , 1 , .1 ) ){
	
	this_ks <-
	    svyafc(
	        ~ wage + collgrad + hours , 
	        design = des_nlsw88 , 
	        cutoffs = list( 4 , 'college grad' , 26 ) , 
	        k = ks , 
	        g = 0 , 
	        na.rm = TRUE 
	       )
	
	page_ten <-
	    rbind(
	        page_ten ,
	        data.frame( 
	            k = ks , 
	            MO = coef( this_ks ) ,
	            seMO = SE( this_ks ) ,
	            H = attr( this_ks , "extra" )[ 1 , 1 ] ,
	            seH = attr( this_ks , "extra" )[ 1 , 2 ] ,
	            A = attr( this_ks , "extra" )[ 2 , 1 ] ,
	            seA = attr( this_ks , "extra" )[ 2 , 2 ]
          )
	    )
	
}

knitr::kable(
  page_ten , caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

still need to replicate PDF page 13

https://github.com/DjalmaPessoa/convey/issues/168


then keep going replicating this

https://github.com/DjalmaPessoa/convey/issues/154





For additional usage examples of `svyafc` or `svyafcdec`, type `?convey::svyafc` or `?convey::svyafcdec` in the R console.



[@alkire2011] and [@alkire2015] and [@pacifico2016]



## Bourguignon (1999) inequality class (svybmi)




For additional usage examples of `svybmi`, type `?convey::svybmi` in the R console.


[@bourguignon1999] and [@lugo2007]

