---
output:
  pdf_document: default
  html_document: default
---
# Multidimensional Indices {#multidimensional}

Inequality and poverty can be seen as multidimensional concepts, combining several livelihood characteristics. Usual approaches take into account income, housing, sanitation, etc. 

In order to transform these different measures from into meaningful numbers, economic theory builds on the idea of utility functions. Utility is a measure of well-being, assigning a "well-being score" to a vector of characteristics. Depending on the utility function, the analyst may allow for substitutions among characteristics: for instance, someone with a slightly lower income, but with access to sanitation, can have a higher wellbeing than someone with a higher income, but without access to sanitation. This depends on the set of weights given to the set of attributes.

Most measures below follow from this kind of two-step procedure: (1) estimating individual scores from an individual's set of characteristics; then (2) aggregating those individual scores into a single measure for the population.

The following section will present measures of multidimensional poverty and inequality, describing the main aspects of the theory and estimation procedures of each.

## Alkire-Foster Class and Decomposition (svyafc, svyafcdec)

This class of measures are defined in @alkire2011, using what is called the "dual cutoff" approach. This method applies a cutoffs to define dimensional deprivations and another cutoff for multidimensional deprivation. 

To analyze a population of $n$ individuals across $d$ achievement dimensions, the first step of the method is applying a FGT-like transformation to each dimension, defined as

\[
g_{ij}^\alpha = \bigg( \frac{ z_j - x_{ij} }{ z_j } \bigg)^{\alpha}
\]

where $i$ is an observation index, $j$ is a dimension index and $\alpha$ is an exponent weighting the deprivation intensity. If $\alpha=0$, then $g_{ij}^0$ becomes a binary variable, assuming value $1$ if person $i$ is deprived in dimension $j$ and $0$ otherwise. The $n \times d$ matrix $G^\alpha$ will be referred to as *deprivation matrix*.

Each dimension receives a weight $w_j$, so that the weighted sum of multidimensional deprivation is the matrix multiplication of $G^\alpha$ by the $j \times 1$ vector $W = [w_j]$. The $n \times 1$ vector $C^\alpha = [c^\alpha_i]$ is the weighted sum of dimensional deprivation scores, i.e.,

\[
c^\alpha_{i} = \sum_{j \in d} w_j g_{ij}^\alpha
\]

The second cutoff is defining those considered to be multidimensionally poor. Assuming that $\sum_{j \in d} w_j = 1$, the multidimensional cutoff $k$ belongs to the interval $(0,1]$. If $c^0_{i} \geqslant k$, then this person is considered multidimensionally poor. The *censored vector of deprivation sums* $C^\alpha(k)$ is defined as

\[
C^\alpha (k) = \bigg[ c_{ij}^\alpha \cdot \delta \big( c_{ij}^0 \geqslant k \big) \bigg] \text{,}
\]

where $\delta(A)$ is an indicator function, taking value $1$ if condition $A$ is true and $0$ otherwise. If $k \geqslant \min{ w_j }$, this is called the "union approach", where a person is considered poor if she is poor in at least one dimension. On the other extreme, the "intersection approach" happens when $k = 1$, meaning that a person is considered poor if she is poor in all dimensions.

The average of vector $C^0 (k)$ returns the multidimensional headcount ratio. For the multidimensional FGT class, a general measure can be defined as

\[
M^\alpha = \frac{1}{n} \sum_{i \in n} \sum_{j \in d} w_j g_{ij}^{\alpha}(k) \text{, } \alpha \geq 0 \text{,}
\]

where $g_{ij}^{\alpha}(k) = g_{ij}^\alpha \cdot \delta \big( c^0_i \geqslant k \big)$.

For inferential purposes, since this variable is actually the average of scores $\sum_{j \in d} w_j g_{ij}^{\alpha}(k)$, the linearization is straightforward.

The Alkire-Foster index is both dimensional and subgroup decomposable. This way, it is possible to analyze how much each dimension or group contribute to the general result. The overall poverty measure can be seen as the weighted sum of each group's poverty measure, as in the formula below:

\[
M^\alpha = \sum_{l \in L} \frac{ n_l }{ n } M^\alpha_{l}
\]

where $l$ is one of $L$ groups.

Also, the overall poverty index can be expressed across dimensions as
\[
M^\alpha = \sum_{j \in d} w_j \bigg[ \frac{1}{n} \sum_{i \in n} g_{ij}^\alpha (k) \bigg] \text{.}
\]

Since those functions are linear combinations of ratios and totals, it is also possible to calculate standard errors for such measures.

---

**A replication example**

In November 2015, Christopher Jindra presented at the Oxford Poverty and Human Development Initiative on the Alkire-Foster multidimensional poverty measure. His presentation can be viewed [here](http://www.ophi.org.uk/wp-content/uploads/Jindra_151109_OPHISeminar.pdf). The example below reproduces those statistics.

Load and prepare the same data set:
```{r message=FALSE, warning=FALSE}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the stata-style webuse library
library(webuse)

# load the same microdata set used by Jindra in his presentation
webuse("nlsw88")

# coerce that `tbl_df` to a standard R `data.frame`
nlsw88 <- data.frame( nlsw88 )

# create a `collgrad` column
nlsw88$collgrad <-
    factor( 
        as.numeric( nlsw88$collgrad ) , 
        label = c( 'not college grad' , 'college grad' ) , 
        ordered = TRUE 
      )

# coerce `married` column to factor
nlsw88$married <- 
    factor( 
        nlsw88$married , 
        levels = 0:1 , 
        labels = c( "single" , "married" ) 
    )

# initiate a linearized survey design object
des_nlsw88 <- svydesign( ids = ~1 , data = nlsw88 )

# immediately run the `convey_prep` function on the survey design
des_nlsw88 <- convey_prep(des_nlsw88)
```


Replicate PDF page 9
```{r message=FALSE, warning=FALSE}
page_nine <-
  svyafc(
    ~ wage + collgrad + hours , 
    design = des_nlsw88 , 
    cutoffs = list( 4, 'college grad' , 26 ) , 
    k = 1/3 , g = 0 , 
    na.rm = TRUE
  )

# MO and seMO
print( page_nine )

# H seH and A seA
print( attr( page_nine , "extra" ) )
```


Replicate PDF page 10
```{r message=FALSE, warning=FALSE}
page_ten <- NULL

# loop through every poverty cutoff `k`
for( ks in seq( 0.1 , 1 , .1 ) ){
	
	this_ks <-
	    svyafc(
	        ~ wage + collgrad + hours , 
	        design = des_nlsw88 , 
	        cutoffs = list( 4 , 'college grad' , 26 ) , 
	        k = ks , 
	        g = 0 , 
	        na.rm = TRUE 
	       )
	
	page_ten <-
	    rbind(
	        page_ten ,
	        data.frame( 
	            k = ks , 
	            MO = coef( this_ks ) ,
	            seMO = SE( this_ks ) ,
	            H = attr( this_ks , "extra" )[ 1 , 1 ] ,
	            seH = attr( this_ks , "extra" )[ 1 , 2 ] ,
	            A = attr( this_ks , "extra" )[ 2 , 1 ] ,
	            seA = attr( this_ks , "extra" )[ 2 , 2 ]
          )
	    )
	
}
```

```{r echo = FALSE}
knitr::kable(
  page_ten , caption = 'PDF Page 10 Replication',
  booktabs = TRUE
)
```


Replicate PDF page 13
```{r}
page_thirteen <- NULL

# loop through every poverty cutoff `k`
for( ks in c( 0.5 , 0.75 , 1 ) ){
	
	this_ks <-
	    svyafc(
        ~ wage + collgrad + hours , 
        design = des_nlsw88 , 
        cutoffs = list( 4, 'college grad' , 26 ) , 
        k = ks , 
        g = 0 , 
        dimw = c( 0.5 , 0.25 , 0.25 ) ,
        na.rm = TRUE
      )
	
	page_thirteen <-
	    rbind(
	        page_thirteen ,
	        data.frame( 
	            k = ks , 
	            MO = coef( this_ks ) ,
	            seMO = SE( this_ks ) ,
	            H = attr( this_ks , "extra" )[ 1 , 1 ] ,
	            seH = attr( this_ks , "extra" )[ 1 , 2 ] ,
	            A = attr( this_ks , "extra" )[ 2 , 1 ] ,
	            seA = attr( this_ks , "extra" )[ 2 , 2 ]
          )
	    )
}
```

```{r echo = FALSE}
knitr::kable(
  page_thirteen , caption = 'PDF Page 13 Replication',
  booktabs = TRUE
)
```

Replicate PDF page 16
```{r}
page_sixteen <- NULL

# loop through every alpha value `g`
for( gs in 0:3 ){
	
	this_gs <-
	    svyafc(
        ~ wage + collgrad + hours , 
        design = des_nlsw88 , 
        cutoffs = list( 4, 'college grad' , 26 ) , 
        k = 1/3 , 
        g = gs , 
        na.rm = TRUE
      )
	
	page_sixteen <-
	    rbind(
	        page_sixteen ,
	        data.frame( 
	            g = gs , 
	            MO = coef( this_gs ) ,
	            seMO = SE( this_gs ) 
          )
	    )
}
```


```{r echo = FALSE}
knitr::kable(
  page_sixteen , caption = 'PDF Page 16 Replication',
  booktabs = TRUE
)
```





Replicate k=1/3 rows of PDF page 17 and 19
```{r}
svyafcdec(
    ~ wage + collgrad + hours , 
    design = des_nlsw88 , 
    cutoffs = list( 4 , 'college grad' , 26 ) , 
    k = 1/3 , 
    g = 0 ,
    na.rm = TRUE
)
```



Replicate PDF pages 21 and 22
```{r}
svyafcdec(
    ~ wage + collgrad + hours , 
    subgroup = ~married , 
    design = des_nlsw88 , 
    cutoffs = list( 4 , 'college grad' , 26 ) , 
    k = 1/3 , 
    g = 0 ,
    na.rm = TRUE
)
```


For additional usage examples of `svyafc` or `svyafcdec`, type `?convey::svyafc` or `?convey::svyafcdec` in the R console.


[@alkire2011] and [@alkire2015] and [@pacifico2016]

## Bourguignon-Chakravarty (2003) multidimensional poverty class

A class of poverty measures is proposed in @bourguignon2003, using a cross-dimensional function that assigns values to each set of dimensionally normalized poverty gaps. It can be defined as:
\[
BCh = \sum_{i \in n} \bigg[ \bigg( \sum_{j \in d} w_{j} x_{ij} \bigg)^{\frac{1}{\theta}} \bigg]^\alpha \text{, } \theta > 0 \text{, } \alpha > 0
\]
where $x_{ij}$ being the normalized poverty gap of dimension $j$ for observation $i$, $w_j$ is the weight of dimension $j$, $\theta$ and $\alpha$ are parameters of the function. 

The parameter $\theta$ is the elasticity of subsitution between the normalized gaps. In another words, $\theta$ defines the order of the weighted generalized mean across achievement dimensions. For instance, when $\theta = 1$, the cross-dimensional aggregation becomes the weighted average of all dimensions. As $\theta$ increases, the importance of the individual's most deprived dimension increases. As @vega2009 points out, it also weights the inequality among deprivations. In its turn, $\alpha$ works as society's poverty-aversion measure parameter. In another words, as $\alpha$ increases, more weight is given to the most deprived individuals. Similar to $\theta$, when $\alpha = 1$, $BCh$ is the average of the weighted deprivation scores. 

## Bourguignon (1999) inequality class (svybmi)

@bourguignon1999 proposes a multidimensional inequality index that possesses interesting properties related to the correlation among the welfare dimensions measured. The estimator used in `convey` comes from the formula presented in @lugo2007 and is defined as:
\[
B_{I} = 1 - \frac{1}{ \widehat{N} } \frac{ \sum_{ i \in S } w_i \bigg[ \sum_{ j \in d } w_j x_{ij} \bigg]^{ \alpha / \beta } }{ 
\bigg[ \sum_{ j \in d } w_j \mu_{ij} \bigg]^{ \alpha / \beta } },
\]
where $\alpha \geqslant 0$ is an inequality-aversion parameter and $\beta \leqslant 1$ is a parameter defining the degree of substitution among dimensions. 

This measure is strong scale-invariant when $\beta = 0$, although @bourguignon1999 demonstrates that strong scale-dependent measures might be interesting in the context of multidimensional inequality. Also, it can be shown that stronger correlation among dimensions leads to less inequality if $\beta > \alpha$.

For additional usage examples of `svybmi`, type `?convey::svybmi` in the R console.
