---
output:
  pdf_document: default
  html_document: default
---



# Future Research

## Multidimensional Poverty

### Example Calculation

## Covariance Matrix

Influence functions and "resampling" replicates can be used to improve the inference 
about differences and changes between estimates. Put simply, the variance of the difference between 
two estimates $\widehat{T}_1$ and $\widehat{T}_2$ is given by

$$
Var \big( \widehat{T}_1 - \widehat{T}_2 \big) = 
    Var \big( \widehat{T}_1 \big) + Var \big( \widehat{T}_2 \big) - 2 Cov \big( \widehat{T}_1 , \widehat{T}_2 \big)
$$

\noindent where: $Var \big( \widehat{T}_1 \big)$ and $Var \big( \widehat{T}_2 \big)$ 
are the variances of the estimators $\widehat{T}$_1 and $\widehat{T}_2$; 
$Cov \big( \widehat{T}_1 , \widehat{T}_2 \big)$ is the covariance of these estimators.
Usually, the covariance term is ignored. But, specially for changes, the covariance
tends to be positive, which results in a conservative variance estimator for the difference.

Then, how can we estimate the covariance? The `survey` package already provides an 
approach for that using the `svyby`function. Based on the `?survey::svyby` examples, 
we have: 

```{r}
# load the survey library
library(survey)

#  load data set
data( api )

# declare sampling design
dclus1 <- svydesign( id=~dnum, weights=~pw, data=apiclus1, fpc=~fpc )

# estimate means
mns <-svyby(~api99, ~stype, dclus1, svymean,covmat=TRUE)

# collect variacnce-covariance matrix of estimates
( m <- vcov( mns ) )

# compute variance terms
var.naive <- sum( diag( m[c(1,3),c(1,3)] ) )
cov.term <- sum( diag( m[ c(1,3),c(3,1)] ) )

# "naive" SE of the difference
sqrt( var.naive )

# SE of the difference
sqrt( var.naive - cov.term )

#... or using svycontrast
svycontrast( mns , c(E = 1, M = -1) )
```

Notice that, because the covariance terms are positive, the (actual) variance of 
the difference is smaller than the "naive" variance.

A similar idea can be implemented with other estimators, such as inequality and poverty measures.
However, this has not yet been implemented for linearization/influence function methods.
In the next section, we show an example with the Gini index.

### Example Calculation

We start by showing what has been implemented --- the resampling based approach:

```{r}
# load the convey package
library(convey)

# load the survey library
library(survey)

# load the laeken library
library(laeken)

# load the synthetic EU statistics on income & living conditions
data(eusilc)

# make all column names lowercase
names(eusilc) <- tolower(names(eusilc))

# construct a survey.design
# using our recommended setup
des_eusilc <-
  svydesign(
    ids = ~ rb030 ,
    strata = ~ db040 ,
    weights = ~ rb050 ,
    data = eusilc
  )

# create bootstrap-based design object
des_eusilc_rep <- as.svrepdesign( des_eusilc , type = "bootstrap" )

# immediately run the convey_prep function on it
des_eusilc <- convey_prep( des_eusilc )
des_eusilc_rep <- convey_prep(des_eusilc_rep)

# estimate gini by gender, with and without covariance matrices
ginis.rep <- svyby( ~ py010n , ~rb090 , des_eusilc_rep , svygini , na.rm = TRUE , covmat = TRUE )
ginis.rep.nocov <- svyby( ~ py010n , ~rb090 , des_eusilc_rep , svygini , na.rm = TRUE , covmat = FALSE )

# variance of the gini difference: naive
svycontrast( ginis.rep.nocov , quote( `male` - `female` ) )

# variance of the gini difference: accounting for covariance
svycontrast( ginis.rep , quote( `male` - `female` ) )
```

While not implemented yet, a linearization-based approach can be applied using:

```{r}
# estimate gini by gender, with and without covariance matrices
ginis.lin.nocov <- svyby( ~ py010n , ~rb090 , des_eusilc , svygini , na.rm = TRUE , covmat = FALSE )
gini.female <- svygini( ~ py010n , subset( des_eusilc , rb090 == "female" ), na.rm = TRUE , influence = TRUE )
gini.male   <- svygini( ~ py010n , subset( des_eusilc , rb090 == "male" ), na.rm = TRUE , influence = TRUE )

# variance of the gini difference: naive
linmat <- rep( 0 , nrow( des_eusilc$variables ) )
linmat[ attr( gini.female , "index" ) ] <- attr( gini.female , "influence" )
linmat[ attr( gini.male , "index" ) ] <- attr( gini.male , "influence" )
linmat <- linmat * des_eusilc$prob
lintot <- svyby( linmat , ~rb090 , des_eusilc , svytotal , covmat = TRUE )

# collect covariance matrix
( m <- vcov( lintot ) )

# SE of the gini difference: naive
SE( svycontrast( ginis.lin.nocov , quote( `male` - `female` ) ) )

# SE of the gini difference: accounting for covariance
SE( svycontrast( lintot , quote( `male` - `female` ) ) )
```
