---
output:
  pdf_document: default
  html_document: default
---



# Multidimensional Poverty

The current `convey` software does not include a measure of multidimensional poverty.  The United Nations Development Program (UNDP) and @alkire2011 at Oxford University's Poverty and Human Development Initiative (OPHI) have published [reports](https://hdr.undp.org/content/2023-global-multidimensional-poverty-index-mpi#/indicies/MPI) and [both Stata and `R` code](https://hdr.undp.org/mpi-statistical-programmes) based on the concepts proposed by Nobel laureate Amartya Sen.  Although we have not incorporated this work into our current `R` package, the example below defines a function and its related decomposition based on the Stata module by @pacifico2016 and then replicates [a presentation by OPHI Research Officer Christop Jindra](https://web.archive.org/web/20181123200434/https://www.ophi.org.uk/wp-content/uploads/Jindra_151109_OPHISeminar.pdf).

This chapter presents two `svyafc` and `svyafcdec` functions as an initial `R` language implementation of multidimensional poverty measurement.  We believe that feedback from users will help us improve this function in future versions of the `R` `convey` package.  We kindly request your comments, criticism, use cases, and suggestions.  We expect that multidimensional poverty measurement will become more popular as researchers focus more attention to this important area.  With that in mind, we look forward to the feedback of readers like you.


## Alkire-Foster Class Function Definition

```{r results='hide', message=FALSE, warning=FALSE}
svyafc <-
  function(formula,
           design,
           k ,
           g ,
           cutoffs ,
           dimw = NULL,
           na.rm = FALSE,
           ...) {
    if (k <= 0 |
        k > 1)
      stop("This functions is only defined for k in (0,1].")
    if (g < 0)
      stop("This function is undefined for g < 0.")
    if (!is.list(cutoffs))
      stop("The parameter 'cutoffs' has to be a list.")
    
    ach.matrix <-
      model.frame(formula, design$variables, na.action = na.pass)[, ]
    
    if (!is.null(dimw)) {
      if (any(is.na(dimw))) {
        stop("Invalid value in dimension weights vector.")
      }
      if (sum(dimw) > 1) {
        stop("The sum of dimension weigths have to be equal to one.")
      }
      if (any(dimw > 1 |
              dimw < 0)) {
        stop("Dim. weights have to be within interval [0,1].")
      }
      if (length(dimw) != ncol(ach.matrix)) {
        stop("Dimension weights' length differs from number of dimensions in formula")
      }
    }
    
    var.class <- lapply(ach.matrix, function(x)
      class(x)[1])
    var.class <- matrix(
      var.class,
      nrow = 1,
      ncol = ncol(ach.matrix),
      dimnames = list(c("var.class"), colnames(ach.matrix))
    )
    
    if (any(!(var.class %in% c("numeric", "integer", "ordered")))) {
      stop(
        "This function is only applicable to variables of types 'numeric' or 'ordered factor'."
      )
    }
    if (any((var.class == "integer"))) {
      stop(
        "At least one of the variables is an integer.\nCoerce your column to numeric with as.numeric if you are sure it's what you want."
      )
    }
    
    w <- 1 / design$prob
    
    if (any(ach.matrix[w != 0, var.class == "numeric"] < 0, na.rm = TRUE))
      stop(
        "The Alkire-Foster multidimensional poverty class is defined for non-negative numeric variables only."
      )
    
    if (na.rm) {
      nas <- apply(ach.matrix, 1, function(x)
        any(is.na(x)))
      design <- design[nas == 0,]
      w <- 1 / design$prob
    }
    
    ach.matrix <-
      model.frame(formula, design$variables, na.action = na.pass)[, ]
    ach.matrix <- ach.matrix [w > 0,]
    w <- w [w > 0]
    
    if (any(is.na(ach.matrix))) {
      if (is.null(dimw)) {
        dimw = rep(1 / ncol(var.class), length(var.class))
      }
      
      rval <- as.numeric(NA)
      variance <- as.numeric(NA)
      class(rval) <- c("cvystat" , "svystat")
      attr(rval, "var") <- variance
      names(rval)[1] <- attr(rval, "statistic") <- "alkire-foster"
      dimtable <-
        as.data.frame(matrix(
          c(strsplit(as.character(formula)[[2]] , ' \\+ ')[[1]], dimw),
          nrow = ncol(var.class),
          ncol = 2,
          dimnames = list(paste("dimension", 1:ncol(var.class)), c("variables", "weight"))
        ),
        stringsAsFactors = FALSE)
      dimtable[, 2] <- as.numeric(dimtable[, 2])
      attr(rval, "dimensions") <- dimtable
      attr(rval, "parameters") <-
        matrix(
          c(g, k),
          nrow = 1,
          ncol = 2,
          dimnames = list("parameters", c("g=", "k="))
        )
      
      return(rval)
      
    }
    
    # Deprivation Matrix
    dep.matrix <-
      sapply(
        seq_along(var.class),
        FUN = function(i) {
          1 * (cutoffs[[i]] > ach.matrix[, i])
        }
      )
    colnames(dep.matrix) <- colnames(var.class)
    
    # Unweighted count of deprivations:
    # depr.count <- rowSums( dep.matrix )
    
    # deprivation k cut
    if (is.null(dimw)) {
      dimw = rep(1 / ncol(dep.matrix), ncol(dep.matrix))
    }
    
    # Weighted sum of deprivations:
    depr.sums <- rowSums(sweep(dep.matrix, MARGIN = 2 , dimw, `*`))
    
    # k multidimensional cutoff:
    multi.cut <- depr.sums * (depr.sums >= k)
    rm(dep.matrix)
    
    # Censored Deprivation Matrix
    cen.dep.matrix <-
      sapply(
        seq_along(cutoffs) ,
        FUN = function(x) {
          if (var.class[x] == "numeric") {
            1 * (cutoffs[[x]] > ach.matrix[, x]) * ((cutoffs[[x]] - ach.matrix[, x]) / cutoffs[[x]]) ^
              g
          } else {
            1 * (cutoffs[[x]] > ach.matrix[, x])
          }
        }
      )
    colnames(cen.dep.matrix) <- colnames(var.class)
    cen.dep.matrix[multi.cut == 0,] <- 0
    
    # Sum of censored deprivations:
    cen.depr.sums <-
      rowSums(sweep(cen.dep.matrix, MARGIN = 2 , dimw, `*`))
    rm(cen.dep.matrix, ach.matrix)
    
    
    w <- 1 / design$prob
    w[w > 0] <- cen.depr.sums
    cen.depr.sums <- w
    rm(w)
    
    if (g == 0) {
      w <- 1 / design$prob
      w[w > 0] <- (depr.sums >= k)
      h_i <- w
      rm(w)
      
      h_est <- survey::svymean(h_i, design)
      
      w <- 1 / design$prob
      w[w > 0] <- multi.cut
      multi.cut <- w
      rm(w)
      
      a_est <- survey::svyratio(multi.cut, h_i, design)
      
      
    }
    
    rval <- survey::svymean(cen.depr.sums , design)
    names(rval)[1] <- attr(rval, "statistic") <- "alkire-foster"
    dimtable <-
      as.data.frame(matrix(
        c(strsplit(as.character(formula)[[2]] , ' \\+ ')[[1]], dimw),
        nrow = ncol(var.class),
        ncol = 2,
        dimnames = list(paste("dimension", 1:ncol(var.class)), c("variables", "weight"))
      ),
      stringsAsFactors = FALSE)
    dimtable[, 2] <- as.numeric(dimtable[, 2])
    attr(rval, "dimensions") <- dimtable
    attr(rval, "parameters") <-
      matrix(
        c(g, k),
        nrow = 1,
        ncol = 2,
        dimnames = list("parameters", c("g=", "k="))
      )
    if (g == 0) {
      attr(rval, "extra") <-
        matrix(
          c(h_est[1], a_est[[1]], attr(h_est, "var")[1] ^ .5, a_est[[2]] ^ .5),
          nrow = 2,
          ncol = 2,
          dimnames = list(c("H", "A"), c("coef", "SE"))
        )
    }
    class(rval) <- c("cvystat" , "svystat")
    
    return(rval)
    
  }
```

## Alkire-Foster Class Decomposition Function Definition


```{r results='hide', message=FALSE, warning=FALSE}
svyafcdec <-
  function(formula,
           subgroup = ~ 1 ,
           design,
           g ,
           cutoffs ,
           k ,
           dimw = NULL,
           na.rm = FALSE,
           ...) {
    if (k <= 0 |
        k > 1)
      stop("This functions is only defined for k in (0,1].")
    if (g < 0)
      stop("This function is undefined for g < 0.")
    if (!is.list(cutoffs))
      stop("The parameter 'cutoffs' has to be a list.")
    if (length(cutoffs) != length(all.vars(formula)))
      stop("number of variables in formula must exactly match cutoffs")
    if (!is.null(dimw)) {
      if (any(is.na(dimw))) {
        stop("Invalid value in dimension weights vector.")
      }
      if (sum(dimw) > 1) {
        stop("The sum of dimension weigths have to be equal to one.")
      }
      if (any(dimw > 1 |
              dimw < 0)) {
        stop("Dim. weights have to be within interval [0,1].")
      }
    }
    
    ach.matrix <-
      model.frame(formula, design$variables, na.action = na.pass)[, ]
    
    if (!is.null(dimw)) {
      if (any(is.na(dimw))) {
        stop("Invalid value in dimension weights vector.")
      }
      if (sum(dimw) > 1) {
        stop("The sum of dimension weigths have to be equal to one.")
      }
      if (any(dimw > 1 |
              dimw < 0)) {
        stop("Dim. weights have to be within interval [0,1].")
      }
      if (length(dimw) != ncol(ach.matrix)) {
        stop("Dimension weights' length differs from number of dimensions in formula")
      }
    }
    
    grpvar <-
      model.frame(subgroup, design$variables, na.action = na.pass)[, ]
    
    var.class <- lapply(ach.matrix, function(x)
      class(x)[1])
    var.class <-
      matrix(
        var.class,
        nrow = 1,
        ncol = ncol(ach.matrix),
        dimnames = list(c("var.class"), colnames(ach.matrix))
      )
    
    if (any(!(var.class %in% c("numeric", "ordered")))) {
      stop(
        "This function is only applicable to variables of types 'numeric' or 'ordered factor'."
      )
    }
    
    w <- 1 / design$prob
    
    if (any(ach.matrix[w != 0, var.class == "numeric"] < 0, na.rm = TRUE))
      stop(
        "The Alkire-Foster multidimensional poverty decompostition is defined for non-negative numeric variables only."
      )
    
    ach.matrix <-
      model.frame(formula, design$variables, na.action = na.pass)[, ]
    grpvar <-
      model.frame(subgroup, design$variables, na.action = na.pass)[, ]
    
    if (class(grpvar) == "labelled") {
      stop("This function does not support 'labelled' variables. Try factor().")
    }
    
    if (na.rm) {
      nas <-
        apply(cbind(ach.matrix, grpvar), 1, function(x)
          any(is.na(x)))
      design <- design[nas == 0,]
    }
    
    w <- 1 / design$prob
    ach.matrix <-
      model.frame(formula, design$variables, na.action = na.pass)[, ]
    grpvar <-
      model.frame(subgroup, design$variables, na.action = na.pass)[, ]
    
    # Deprivation Matrix
    dep.matrix <-
      sapply(
        seq_along(var.class),
        FUN = function(i) {
          1 * (cutoffs[[i]] > ach.matrix[, i])
        }
      )
    colnames(dep.matrix) <- colnames(var.class)
    
    # Unweighted count of deprivations:
    # depr.count <- rowSums( dep.matrix )
    
    # deprivation k cut
    if (is.null(dimw)) {
      dimw = rep(1 / ncol(dep.matrix), ncol(dep.matrix))
    }
    
    # Weighted sum of deprivations:
    depr.sums <- rowSums(sweep(dep.matrix, MARGIN = 2 , dimw, `*`))
    
    # k multidimensional cutoff:
    multi.cut <- depr.sums * (depr.sums >= k)
    #rm(dep.matrix)
    
    # Censored Deprivation Matrix
    cen.dep.matrix <-
      sapply(
        seq_along(cutoffs) ,
        FUN = function(x) {
          if (var.class[x] == "numeric") {
            1 * (cutoffs[[x]] > ach.matrix[, x]) * ((cutoffs[[x]] - ach.matrix[, x]) / cutoffs[[x]]) ^
              g
          } else {
            1 * (cutoffs[[x]] > ach.matrix[, x])
          }
        }
      )
    colnames(cen.dep.matrix) <- colnames(var.class)
    
    cen.dep.matrix[multi.cut == 0 & !is.na(multi.cut),] <- 0
    
    # Sum of censored deprivations:
    cen.depr.sums <-
      rowSums(sweep(cen.dep.matrix, MARGIN = 2 , dimw, `*`))
    
    if (any(is.na(cen.depr.sums)[w > 0])) {
      # overall result:
      overall.result <-
        matrix(
          c(NA, NA),
          nrow = 1,
          ncol = 2,
          dimnames = list("overall", c("alkire-foster", "SE"))
        )
      
      # group breakdown:
      if (!is.null(levels(grpvar))) {
        grp.estim <-
          matrix(
            rep(NA, 2 * length(levels(grpvar))),
            nrow = length(levels(grpvar)),
            ncol = 2,
            dimnames = list(levels(grpvar), c("alkire-foster", "SE"))
          )
        grp.contr <-
          matrix(
            rep(NA, 2 * length(levels(grpvar))),
            nrow = length(levels(grpvar)),
            ncol = 2,
            dimnames = list(levels(grpvar), c("contribution", "SE"))
          )
        
      }
      
      # dimensional breakdown:
      dim.raw.hc <-
        matrix(
          rep(NA, 2 * ncol(cen.dep.matrix)),
          nrow = ncol(cen.dep.matrix),
          ncol = 2,
          dimnames = list(colnames(cen.dep.matrix), c("raw headcount", "SE"))
        )
      dim.cen.hc <-
        matrix(
          rep(NA, 2 * ncol(cen.dep.matrix)),
          nrow = ncol(cen.dep.matrix),
          ncol = 2,
          dimnames = list(colnames(cen.dep.matrix), c("cens. headcount", "SE"))
        )
      dim.contri <-
        matrix(
          rep(NA, 2 * ncol(cen.dep.matrix)),
          nrow = ncol(cen.dep.matrix),
          ncol = 2,
          dimnames = list(colnames(cen.dep.matrix), c("contribution", "SE"))
        )
      
      # set up result object:
      if (is.null(levels(grpvar))) {
        rval <- list(overall.result, dim.raw.hc, dim.cen.hc, dim.contri)
      } else {
        rval <-
          list(overall.result,
               dim.raw.hc,
               dim.cen.hc,
               dim.contri,
               grp.estim,
               grp.contr)
      }
      
      return(rval)
      
    }
    
    raw.hc.ratio <-
      survey::svymean(dep.matrix[, ] , design, na.rm = TRUE)
    attr(raw.hc.ratio, "statistic") <- "raw headcount"
    cen.hc.ratio <-
      survey::svymean(cen.dep.matrix[, ] , design, na.rm = TRUE)
    attr(cen.hc.ratio, "statistic") <- "cens. headcount"
    
    U_0 <-
      list(value = sum(w[w > 0]), lin = rep(1, length(w)))
    U_1 <-
      list(value = sum(w[w > 0] * cen.depr.sums[w > 0]), lin = cen.depr.sums)
    
    # overall alkire-foster index:
    overall <- survey::svymean(cen.depr.sums, design, na.rm = TRUE)
    names(overall)[1] <-
      attr(overall, "statistic") <- "alkire-foster"
    
    # group decomposition
    if (!is.null(levels(grpvar))) {
      grp.pctg.estm <- NULL
      grp.pctg.estm_var <- NULL
      grp.pctg.cont <- NULL
      grp.pctg.cont_lin <-
        matrix(data = rep(w, length(levels(grpvar))), nrow = length(w))
      for (i in seq_along(levels(grpvar))) {
        w_i <- w * (grpvar == levels(grpvar)[i])
        U_1_i <-
          list(value = sum(cen.depr.sums[w_i > 0] * w_i[w_i > 0]),
               lin = cen.depr.sums * (grpvar == levels(grpvar)[i]))
        U_0_i <-
          list(value = sum(w_i[w_i > 0]),
               lin = 1 * (grpvar == levels(grpvar)[i]))
        
        list_all <-
          list(
            U_0 = U_0,
            U_1 = U_1,
            U_0_i = U_0_i,
            U_1_i = U_1_i
          )
        
        estimate <-
          contrastinf(quote((U_1_i / U_0_i)), list_all)
        grp.pctg.estm[i] <- estimate$value
        estimate$lin[is.na(estimate$lin)] <- 0
        grp.pctg.estm_var[i] <-
          survey::svyrecvar(
            estimate$lin * w_i,
            design$cluster,
            design$strata,
            design$fpc,
            postStrata = design$postStrata
          )
        
        estimate <-
          contrastinf(quote((U_0_i / U_0) * (U_1_i / U_0_i) / (U_1 / U_0)), list_all)
        grp.pctg.cont[i] <- estimate$value
        estimate$lin[is.na(estimate$lin)] <- 0
        grp.pctg.cont_lin[, i] <- estimate$lin
        
      }
      
      grp.pctg.cont_var <-
        survey::svyrecvar(
          grp.pctg.cont_lin * w,
          design$cluster,
          design$strata,
          design$fpc,
          postStrata = design$postStrata
        )
      
      grp.contr.estimate <-
        matrix(grp.pctg.estm,
               ncol = 1,
               dimnames = list(levels(grpvar), "alkire-foster"))
      attr(grp.contr.estimate, "names") <- levels(grpvar)
      attr(grp.contr.estimate, "var") <- grp.pctg.estm_var
      attr(grp.contr.estimate, "statistic") <- "alkire-foster"
      class(grp.contr.estimate) <- c("svystat")
      
      grp.contr.pct <-
        matrix(
          grp.pctg.cont,
          ncol = 1,
          dimnames = list(levels(grpvar), "grp. % contribution")
        )
      attr(grp.contr.pct, "names") <- levels(grpvar)
      attr(grp.contr.pct, "var") <- grp.pctg.cont_var
      attr(grp.contr.pct, "statistic") <- "grp. % contribution"
      class(grp.contr.pct) <- c("svystat")
      
      rm(
        grp.pctg.estm,
        grp.pctg.estm_var,
        grp.pctg.cont,
        grp.pctg.cont_lin,
        grp.pctg.cont_var
      )
      
    }
    
    # dimensional decomposition:
    dim.contr <- NULL
    dim.contr_lin <-
      matrix(data = rep(w, ncol(cen.dep.matrix)), nrow = length(w))
    for (i in 1:ncol(cen.dep.matrix)) {
      wj <-
        list(value = dimw[i], lin = rep(0, length(cen.depr.sums)))
      
      H_0 <-
        list(value = sum(cen.dep.matrix[w > 0 , i] * w[w > 0]),
             lin = cen.dep.matrix[, i] * (w > 0))
      list_all <- list(
        U_0 = U_0,
        U_1 = U_1,
        H_0 = H_0,
        wj = wj
      )
      estimate <-
        contrastinf(quote(wj * (H_0 / U_0) / (U_1 / U_0)), list_all)
      
      dim.contr[i] <- estimate$value
      estimate$lin[is.na(estimate$lin)] <- 0
      dim.contr_lin[, i] <- estimate$lin
      
    }
    
    dim.contr_var <-
      survey::svyrecvar(
        dim.contr_lin * w,
        design$cluster,
        design$strata,
        design$fpc,
        postStrata = design$postStrata
      )
    
    dim.result <- dim.contr
    attr(dim.result, "names") <- colnames(ach.matrix)
    attr(dim.result, "var") <- dim.contr_var
    attr(dim.result, "statistic") <- "dim. % contribution"
    class(dim.result) <- c("svystat")
    
    rm(dim.contr, dim.contr_lin, dim.contr_var)
    
    # set up result object:
    if (is.null(levels(grpvar))) {
      rval <- list(overall, raw.hc.ratio, cen.hc.ratio, dim.result)
      names(rval) <-
        list(
          "overall",
          "raw headcount ratio",
          "censored headcount ratio",
          "percentual contribution per dimension"
        )
      
    } else {
      rval <-
        list(
          overall,
          raw.hc.ratio,
          cen.hc.ratio,
          dim.result,
          grp.contr.estimate,
          grp.contr.pct
        )
      names(rval) <-
        list(
          "overall",
          "raw headcount ratio",
          "censored headcount ratio",
          "percentual contribution per dimension",
          "subgroup alkire-foster estimates",
          "percentual contribution per subgroup"
        )
      
    }
    
    return(rval)
    
  }
```

## Replication Example

The `R` code below replicates multiple statistics and standard errors from [the presentation by OPHI Research Officer Christop Jindra](https://web.archive.org/web/20181123200434/https://www.ophi.org.uk/wp-content/uploads/Jindra_151109_OPHISeminar.pdf):


```{r}
library(survey)
library(convey)
library(webuse)


# load the same microdata set used by Jindra in his presentation
webuse("nlsw88")

# coerce that `tbl_df` to a standard R `data.frame`
nlsw88 <- data.frame(nlsw88)

# create a `collgrad` column
nlsw88$collgrad <-
  factor(
    as.numeric(nlsw88$collgrad) ,
    label = c('not college grad' , 'college grad') ,
    ordered = TRUE
  )

# initiate a linearization-based survey design object
des_nlsw88 <- svydesign(ids = ~ 1 , data = nlsw88)

# immediately run the `convey_prep` function on the survey design
des_nlsw88 <- convey_prep(des_nlsw88)

# PDF page 9
result <-
  svyafc(
    ~ wage + collgrad + hours,
    design = des_nlsw88,
    cutoffs = list(4, 'college grad' , 26),
    k = 1 / 3,
    g = 0,
    na.rm = T
  )

result

attr(result, 'extra')


# PDF page 10
for (ks in seq(0.1, 1, .1)) {
  result <-
    svyafc(
      ~ wage + collgrad + hours,
      design = des_nlsw88,
      cutoffs = list(4, 'college grad' , 26),
      k = ks ,
      g = 0,
      na.rm = T
    )
  
  print(result)
  
  print(attr(result , 'extra'))
  
}

# PDF page 13
for (ks in c(0.5 , 0.75 , 1)) {
  result <-
    svyafc(
      ~ wage + collgrad + hours ,
      design = des_nlsw88 ,
      cutoffs = list(4, 'college grad' , 26) ,
      k = ks ,
      g = 0 ,
      dimw = c(0.5 , 0.25 , 0.25) ,
      na.rm = TRUE
    )
  
  print(result)
  
  print(attr(result, 'extra'))
  
}

# PDF page 17
result <-
  svyafc(
    ~ wage + collgrad + hours,
    design = des_nlsw88,
    cutoffs = list(4, 'college grad' , 26),
    k = 1 / 3,
    g = 0,
    na.rm = T
  )

result

attr(result, "extra")

svyafcdec(
  ~ wage + collgrad + hours,
  by = ~ 1,
  design = des_nlsw88,
  cutoffs = list(4, 'college grad' , 26),
  k = 1 / 3,
  g = 0,
  na.rm = T
)[2]

# PDF page 19

# Censored headcount ratios
for (ks in seq(1 / 3, 1, 1 / 3)) {
  print (
    svyafcdec(
      ~ wage + collgrad + hours,
      by = ~ 1,
      design = des_nlsw88,
      cutoffs = list(4, 'college grad' , 26),
      k = ks,
      g = 0,
      na.rm = T
    )[3]
  )
  
}
# Percentage contribution by dimensions
for (ks in seq(1 / 3, 1, 1 / 3)) {
  print (
    svyafcdec(
      ~ wage + collgrad + hours,
      by = ~ 1,
      design = des_nlsw88,
      cutoffs = list(4, 'college grad' , 26),
      k = ks,
      g = 0,
      na.rm = T
    )[4]
  )
  
}

# PDF page 22
for (ks in seq(1 / 3, 1, 1 / 3)) {
  print(
    svyafcdec(
      ~ wage + collgrad + hours,
      subgroup = ~ factor(married),
      design = des_nlsw88,
      cutoffs = list(4, 'college grad' , 26),
      k = ks ,
      g = 0,
      dimw = NULL,
      na.rm = T
    )[6]
  )
  
}
```

