--- 
title: "Poverty and Inequality with Complex Survey Data"
author: "Guilherme Jacob, Anthony Damico, and Djalma Pessoa"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site

output:
  bookdown::tufte_html_book:
    toc: yes
    css: toc.css
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    latex_engine: xelatex
    citation_package: natbib
    keep_tex: yes
  bookdown::epub_book: default

    
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: gilhermejacob/context
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

# Introduction

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

For now, you have to install the development version of **bookdown** from Github:
```{r eval=FALSE}
devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need to install XeLaTeX.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```



The library convey aims at estimating measures of poverty and income concentration. There are already at least two libraries covering this subject: vardpoor and Laeken.
The main difference between the library convey and these two is that the convey strongly hinges on the survey library.


## Installation {#install}

`convey` is free and open-source software that runs inside the [R environment for statistical computing](https://www.r-project.org/).

* the latest released version from [CRAN](https://CRAN.R-project.org/package=convey) with

    ```R
    install.packages("convey")
    ````

* the latest development version from github with

    ```R
    devtools::install_github("djalmapessoa/convey")
    ```


[This may present how to install R, RStudio and required packages. Providing brief information about `survey` and `MonetDBLite` may also be recommended.]

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(install). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(inequality).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

## Complex surveys and statistical inference {#survey}

In this book we estimate measures of poverty and income concentration in a populatiion, generaly of households or people, based on data colected from a complex survey sample from the population, involving

1- different units selection probabilities;

2- clustering of units;

3- stratification of clusters, and

4- reweighting to compensate missing values and other adjustments.

Items 1 and 4 imply that we should use different units weights to avoid biases when performing statistical analysis. Also, when estimating variances, we should consider, not only the design weights but all listed design characteristics 1-4.

In order to take into account the sample design characteristics it should be used a specialized software like the R library **survey**, adopted in this book.



## Linearization

Some measures of poverty and income concentration are defined by non-differentiable functions so that it is not possible to use Taylor linearization to estimate their variances. An alternative is to use **Influence functions** as described in [@deville1999] and [@osier2009]. The library convey implements this methodology to work with `survey.design` objects and also with `svyrep.design` objects.

Some examples of these measures are:

- At-risk-of-poverty threshold:
$arpt=.60q_{.50}$ where $q_{.50}$ is the income median;

- At-risk-of-poverty rate
$arpr=\frac{\sum_U 1(y_i \leq arpt)}{N}.100$

- Quintile share ratio
     
$qsr=\frac{\sum_U 1(y_i>q_{.80})}{\sum_U 1(y_i\leq q_{.20})}$

- Gini coefficient
$1+G=\frac{2\sum_U (r_i-1)y_i}{N\sum_Uy_i}$
where $r_i$ is the rank of $y_i$.

Note that it is not possible to use Taylor linearization for these measures because they depend on quantiles and the Gini is defined as a function of ranks. This could be done using the approach proposed by Deville (1999) based upon influence functions. 



## Influence function


Let $U$ be a population of size $N$ and $M$ be a measure that allocates mass one to the set composed by one unit, that is $M(i)=M_i= 1$ if $i\in U$ and $M(i)=0$ if $i\notin U$ 

Now, a population parameter $\theta$ can be expressed as a functional of $M$
$\theta=T(M)$

Examples of such parameters are:

- Total: 
$Y=\sum_Uy_i=\sum_U y_iM_i=\int ydM=T(M)$

- Ratio of two totals:
$R=\frac{Y}{X}=\frac{\int y dM}{\int x dM}=T(M)$

- Cumulative distribution function:
$F(x)=\frac{\sum_U 1(y_i\leq x)}{N}=\frac{\int 1(y\leq x)dM}{\int{dM}}=T(M)$


To estimate these parameters from the sample, we replace the measure $M$ by the estimated measure $\hat{M}$ defined by: $\hat{M}(i)=\hat{M}_i= w_i$ if $i\in s$ and $\hat{M}(i)=0$ if $i\notin s$. 

The estimators of the population parameters can then be expressed as functional of the measure  $\hat{M}$. 

-  Total:
$\hat{Y}=T(\hat{M})=\int yd\hat{M}=\sum_s w_iy_i$

- Ratio of totals:
$\hat{R}=T(\hat{M})=\frac{\int y d\hat{M}}{\int x d\hat{M}}=\frac{\sum_s w_iy_i}{\sum_s w_ix_i}$

- Cumulative distribution function:
$\hat{F}(x)=T(\hat{M})=\frac{\int 1(y\leq x)d\hat{M}}{\int{d\hat{M}}}=\frac{\sum_s w_i 1(y_i\leq x)}{\sum_s w_i}$



## The variance estimator

The variance of the estimator $T(\hat{M})$ can approximated by:

$$Var\left[T(\hat{M})\right]\cong var\left[\sum_s w_i z_i\right]$$

The  `linearized` variable $z$  is given by the derivative of the functional:

$$
z_k=lim_{t\rightarrow0}\frac{T(M+t\delta_k)-T(M)}{t}=IT_k(M)
$$

where, $\delta_k$ is the Dirac measure in $k$: $\delta_k(i)=1$ if and only if $i=k$.

This **derivative** is called  **Influence Function** and was introduced in the area of **Robust Statistics**.

## Influence functions - Examples


- Total:
$$
\begin{aligned}
IT_k(M)&=lim_{t\rightarrow 0}\frac{T(M+t\delta_k)-T(M)}{t}\\
&=lim_{t\rightarrow 0}\frac{\int y.d(M+t\delta_k)-\int y.dM}{t}\\
&=lim_{t\rightarrow 0}\frac{\int yd(t\delta_k)}{t}=y_k	
\end{aligned}
$$


- Ratio of two totals:
$$
\begin{aligned}
IR_k(M)&=I\left(\frac{U}{V}\right)_k(M)=\frac{V(M)\times IU_k(M)-U(M)\times IV_k(M)}{V(M)^2}\\
&=\frac{X y_k-Y x_k}{X^2}=\frac{1}{X}(y_k-Rx_k)
\end{aligned}
$$

## Linearization by influence function - Examples

- At-risk-of-poverty threshold:
$$
arpt = 0.6\times m
$$
where $m$ is the median income.

$$
z_k= -\frac{0.6}{f(m)}\times\frac{1}{N}\times\left[I(y_k\leq m-0.5) \right]
$$


- At-risk-of-poverty rate:

\[
 arpr=\frac{\sum_U I(y_i \leq t)}{\sum_U w_i}.100
\]
\[
z_k=\frac{1}{N}\left[I(y_k\leq t)-t\right]-\frac{0.6}{N}\times\frac{f(t)}{f(m)}\left[I(y_k\leq m)-0.5\right]
\]

where:

$N$ - population size; 

$t$ - at-risk-of-poverty threshold;

$y_k$ - income of person $k$;

$m$ - median income;

$f$ - income density function;

## Structure of the library

In the library convey, there are some basic functions that produces the linearized variables of some estimates that often enter in the definition of measures of concentration and poverty. For example the `quantile` which is linearized by the function `svyiqalpha`. Other example is the function `svyisq` that linearizes the total below a quantile of the variable.

From the linearized variables of these basic estimates it is possible by using rules of composition, valid for influence functions, to derive the influence function of more complex estimates. By definition the influence function is a Gateaux derivative and the rules rules of composition valid for Gateaux derivatives also hold for Influence Functions.

The following property of Gateaux derivatives was often used in the library convey. Let $g$ be a differentible function of $m$ variables. Suppose we want to compute the influence function of the estimator $g(T_1, T_2,\ldots, T_m)$, knowing the Influence function of the estimators $T_i, i=1,\ldots, m$. Then the following holds:

$$
I(g(T_1, T_2,\ldots, T_m)) = \sum_{i=1}^m \frac{\partial g}{\partial T_i}I(T_i)
$$

In the library convey this rule is implemented by the function `contrastinf` which uses the R function `deriv` to compute the formal partial derivatives $\frac{\partial g}{\partial T_i}$. 

For example, suppose we want to linearize the `Relative median poverty gap`(rmpg), defined as the difference between the at-risk-of-poverty threshold (`arpt`) and the median of incomes less than the `arpt` relative to the `arprt`: 

$$
rmpg= \frac{arpt-medpoor} {arpt}
$$

where `medpoor` is the median of incomes less than `arpt`.

Suppose we know how to linearize `arpt` and `medpoor`, then by applying the function `contrastinf` with 
$$
g(T_1,T_2)= \frac{(T_1 - T_2)}{T_1}
$$
we linearize the `rmpg`.


## Examples of use of the library convey


In the following examples we will use the data set `eusilc` contained in the libraries `vardpoor` and `Laeken`.

```{r results='hide', message=FALSE, warning=FALSE}
library(vardpoor)
data(eusilc)
```
Next, we create an object of class `survey.design` using the function `svydesign` of the library survey:

```{r results='hide', message=FALSE, warning=FALSE}
library(survey)
des_eusilc <- svydesign(ids = ~rb030, strata =~db040,  weights = ~rb050, data = eusilc)
```
Right after the creation of the design object `des_eusilc`, we should use the function `convey_prep` that adds an attribute to the survey design which saves information on the design object based upon the whole sample, needed to work with subset designs.

```{r}
library(convey)
des_eusilc <- convey_prep( des_eusilc )
```
To estimate the `at-risk-of-poverty rate` we use the function `svyarpt`:

```{r comment=NA}
svyarpr(~eqIncome, design=des_eusilc)
```
To estimate the `at-risk-of-poverty rate` for domains defined by the variable `db040` we use

```{r comment=NA}
svyby(~eqIncome, by = ~db040, design = des_eusilc, FUN = svyarpr, deff = FALSE)
```

Using the same data set, we estimate the `quintile share ratio`: 

```{r comment=NA}
# for the whole population
svyqsr(~eqIncome, design=des_eusilc, alpha= .20)

# for domains
svyby(~eqIncome, by = ~db040, design = des_eusilc,
  FUN = svyqsr, alpha= .20, deff = FALSE)

```

These functions can be used as S3 methods for the classes `survey.design` and `svyrep.design`.

Let's create a design object of class `svyrep.design` and run the function `convey_prep` on it:

```{r}
des_eusilc_rep <- as.svrepdesign(des_eusilc, type = "bootstrap")
des_eusilc_rep <- convey_prep(des_eusilc_rep) 
```

and then use the function `svyarpr`:

```{r comment=NA}
svyarpr(~eqIncome, design=des_eusilc_rep)
svyby(~eqIncome, by = ~db040, design = des_eusilc_rep, FUN = svyarpr, deff = FALSE)
```
The functions of the library convey are called in a similar way  to the functions in library survey.

It is also possible to deal with missing values by using the argument `na.rm`.

```{r comment=NA}
# survey.design using a variable with missings
svygini( ~ py010n , design = des_eusilc )
svygini( ~ py010n , design = des_eusilc , na.rm = TRUE )
# svyrep.design using a variable with missings
# svygini( ~ py010n , design = des_eusilc_rep ) get error
svygini( ~ py010n , design = des_eusilc_rep , na.rm = TRUE )
```



djalma, where do these references go on this page?  [@berger2003] and [@osier2009] and [@deville1999]


