--- 
title: "Poverty and Inequality with Complex Survey Data"
author: "Guilherme Jacob, Anthony Damico, and Djalma Pessoa"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::tufte_html_book:
    toc: yes
    css: toc.css
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    latex_engine: xelatex
    citation_package: biblatex
    keep_tex: yes
  bookdown::epub_book: default
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apa
link-citations: yes
github-repo: guilhermejacob/context
description: "A book about the R convey package"
delete_merged_file: yes
---

```{r results='hide', echo=FALSE}
set.seed(2017)
```

# Introduction

The R `convey` library estimates measures of poverty, inequality, and wellbeing.  There are two other R libraries covering this subject, [vardpoor](https://CRAN.R-project.org/package=vardpoor) [@R-vardpoor] and [laeken](https://CRAN.R-project.org/package=laeken) [@R-laeken], however, only `convey` integrates seamlessly with the [R survey package](https://CRAN.R-project.org/package=survey) [@R-survey-article;@R-survey-book;@R-survey].

`convey` is free and open-source software that runs inside the [R environment for statistical computing](https://www.r-project.org/).  Anyone can review and propose changes to [the source code](https://github.com/ajdamico/convey) for this software.  Readers are welcome to [propose changes to this book](https://github.com/guilhermejacob/context/) as well.

Individuals getting started in the field of poverty and inequality statistics might find the number of techniques described in this textbook overwhelming, especially on the topic of choosing which method might be most appropriate for each particular research question.  For this reason, the authors of this textbook consider Dr. Ija Trapeznikova's article [Measuring income inequality](https://wol.iza.org/articles/measuring-income-inequality/long) an important summary of how to approach selecting between available techniques.


## Installation {#install}

In order to work with the `convey` library, you will need to have R running on your machine.  If you have never used R before, you will need to [install that software](https://www.r-project.org/) before `convey` can be accessed.  Once you have R loaded on your machine, you can install..

* the latest released version from [CRAN](https://CRAN.R-project.org/package=convey) with

```R
install.packages("convey")
```

* the latest development version from github with

```R
remotes::install_github("ajdamico/convey")
```

In order to know how to cite this package, run `citation("convey")`.

## Complex surveys and statistical inference {#survey}

In this book, we demonstrate how to measure poverty and income concentration in a population based on microdata collected from a complex survey sample.  Most surveys administered by government agencies or larger research organizations utilize a sampling design that violates the assumption of simple random sampling (SRS), including:

1. Different units selection probabilities;
2. Clustering of units;
3. Stratification of clusters;
4. Reweighting to compensate for missing values and other adjustments.

Therefore, basic unweighted R commands such as `mean()` or `glm()` will not properly account for the weighting nor the measures of uncertainty (such as the confidence intervals) present in the dataset.  For some examples of publicly-available complex survey data sets, see [http://asdfree.com]().  

Unlike other software, the R `convey` package does not require that the user specify these parameters throughout the analysis.  So long as the [svydesign object](http://r-survey.r-forge.r-project.org/survey/html/svydesign.html) or [svrepdesign object](http://r-survey.r-forge.r-project.org/survey/html/svrepdesign.html) has been constructed properly at the outset of the analysis, the `convey` package will incorporate the survey design automatically and produce statistics and variances that take the complex sample into account.

Survey analysts familiar with the R `dplyr` syntax implemented by the `survey` library's wrapper `srvyr` package might be interested in implementing specific `convey` functions by following the [`svygini()` example](http://gdfe.co/srvyr/articles/extending-srvyr.html) published by `srvyr` author Greg Freedman Ellis.  Note that the full design stored by `convey_prep()` may in some cases complicate this extension.


## Usage Examples


In the following example, we've loaded the data set `eusilc` from the R library [laeken](https://CRAN.R-project.org/package=laeken) [@R-laeken].

```{r results='hide', message=FALSE, warning=FALSE}
library(laeken)
data(eusilc)
```
Next, we create an object of class `survey.design` using the function `svydesign` of the `survey` library:

```{r results='hide', message=FALSE, warning=FALSE}
library(survey)
des_eusilc <-
  svydesign(
    ids = ~ rb030,
    strata =  ~ db040,
    weights = ~ rb050,
    data = eusilc
  )
```
Right after the creation of the design object `des_eusilc`, we should use the function `convey_prep` that adds an attribute to the survey design which saves information on the design object based upon the whole sample, needed to work with subset designs.

```{r}
library(convey)
des_eusilc <- convey_prep(des_eusilc)
```
To estimate the at-risk-of-poverty rate, we use the function `svyarpt`:

```{r comment=NA}
svyarpr(~ eqIncome, design = des_eusilc)
```
To estimate the at-risk-of-poverty rate across domains defined by the variable `db040` we use:

```{r comment=NA}
svyby(
  ~ eqIncome,
  by = ~ db040,
  design = des_eusilc,
  FUN = svyarpr,
  deff = FALSE
)
```

Using the same data set, we estimate the quintile share ratio: 

```{r comment=NA}
# for the whole population
svyqsr( ~ eqIncome, design = des_eusilc, alpha1 = .20)

# for domains
svyby(
  ~ eqIncome,
  by = ~ db040,
  design = des_eusilc,
  FUN = svyqsr,
  alpha1 = .20,
  deff = FALSE
)
```

These functions can be used as S3 methods for the classes `survey.design` and `svyrep.design`.

Let's create a design object of class `svyrep.design` and run the function `convey_prep` on it:

```{r}
des_eusilc_rep <- as.svrepdesign(des_eusilc, type = "bootstrap")
des_eusilc_rep <- convey_prep(des_eusilc_rep)
```

The function `svyarpr` produces matching coefficients and near-identical standard errors on the replication design:

```{r comment=NA}
svyarpr( ~ eqIncome, design = des_eusilc_rep)

svyby(
  ~ eqIncome,
  by = ~ db040,
  design = des_eusilc_rep,
  FUN = svyarpr,
  deff = FALSE
)
```
The functions of the convey `library` are called in a similar way  to the functions in `survey` library.

It is also possible to deal with missing values by using the argument `na.rm`.

```{r comment=NA}
# survey.design using a variable with missings
svygini( ~ py010n , design = des_eusilc )
svygini( ~ py010n , design = des_eusilc , na.rm = TRUE )

# svyrep.design using a variable with missings
svygini( ~ py010n , design = des_eusilc_rep )
svygini( ~ py010n , design = des_eusilc_rep , na.rm = TRUE )
```



## Underlying Calculations {#lin}

In what follows, we often use the linearization method as a tool to produce an approximation for the variance of an estimator. From the linearized variable $z$ of an estimator $T$, we get from the expression \@ref(var) an estimate of the variance of $T$.

If $T$ can be expressed as a function of the population totals $T = g(Y_1, Y_2, \ldots, Y_n)$, and if $g$ is linear,  the estimation of the variance of $T = g(Y_1, Y_2, \ldots, Y_n)$ is straightforward. If $g$ is not linear but is a 'smooth' function, then it is possible to approximate the variance of $g(Y_1, Y_2, \ldots, Y_n)$ by the variance of its first order Taylor expansion. For example, we can use Taylor expansion to linearize the ratio of two totals. However, there are situations where Taylor linearization cannot be immediately computed, either because $T$ cannot be expressed as functions of the population totals, or because $g$ is not  a `smooth` function.  A common example is the case where $T$ is a quantile.

In these cases, an alternative form of linearization of $T$ might suffice: the `Influence Function`, as defined in \@ref(lin), proposed in @deville1999.  Separately, replication methods such as `bootstrap` and `jackknife` also work.

In the `convey` library, there are some basic functions that produce the linearized variables needed to measure income concentration and poverty.  For example, looking at the income variable in some complex survey dataset, the `quantile` of that income variable can be linearized by the function `convey::svyiqalpha` and the sum total below any quantile of the variable is linearized by the function `convey::svyisq`.

From the linearized variables of these basic estimates, it is possible by using rules of composition, valid for influence functions, to derive the influence function of more complex estimates. By definition, the influence function is a Gateaux derivative and the rules of composition valid for Gateaux derivatives also hold for Influence Functions.

The following property of Gateaux derivatives is commonly used in the `convey` library:  Let $g$ be a differentiable function of $m$ variables. Suppose we want to compute the influence function of the estimator $g(T_1, T_2,\ldots, T_m)$, knowing the influence function of the estimators $T_i, i=1,\ldots, m$. Then the following holds:

$$
I(g(T_1, T_2,\ldots, T_m)) = \sum_{i=1}^m \frac{\partial g}{\partial T_i}I(T_i)
$$

In the `convey` library, this rule is implemented by the function `contrastinf`, which uses the base R function `deriv` to compute the formal partial derivatives $\frac{\partial g}{\partial T_i}$. 

For example, suppose we want to linearize the relative median poverty gap (RMPG), defined as the difference between the at-risk-of-poverty threshold (ARPT) and the median of incomes less than the ARPT, relative to the ARPT itself.  Let's say that this median income below the at-risk-of-poverty-threshold (POORMED) is the median of incomes less than ARPT:


$$
rmpg= \frac{(arpt-poormed)} {arpt}
$$


If we know how to linearize ARPT and POORMED, then by applying the function `contrastinf` with 
$$
g(T_1,T_2)= \frac{(T_1 - T_2)}{T_1}
$$
we are also able to linearize the RMPG.



## The Variance Estimator {#var}


Using the notation in @osier2009, the variance of the estimator $T(\hat{M})$ can approximated by:

\begin{equation}
Var\left[T(\hat{M})\right]\cong var\left[\sum_s w_i z_i\right]
(\#var)
\end{equation}

The  `linearized` variable $z$  is given by the derivative of the functional:

\begin{equation}
z_k=lim_{t\rightarrow0}\frac{T(M+t\delta_k)-T(M)}{t}=IT_k(M)
(\#lin)
\end{equation}

where, $\delta_k$ is the Dirac measure in $k$: $\delta_k(i)=1$ if and only if $i=k$.

This **derivative** is called the **Influence Function** and was introduced in the area of **Robust Statistics**. 



## Influence Functions

Some measures of poverty and income concentration are defined by non-differentiable functions, so that it is not always possible to use Taylor linearization to estimate variances. An alternative is to use **influence functions** as described in @deville1999 and @osier2009. The `convey` library implements this methodology to work with `survey.design` objects and also with `svyrep.design` objects.

Some examples of these measures are:

- At-risk-of-poverty threshold:
$arpt=.60q_{.50}$ where $q_{.50}$ is the median income;

- At-risk-of-poverty rate:
$arpr=\frac{\sum_U 1(y_i \leq arpt)}{N}.100$

- Quintile share ratio:
$qsr=\frac{\sum_U 1(y_i>q_{.80})}{\sum_U 1(y_i\leq q_{.20})}$

- Gini coefficient
$1+G=\frac{2\sum_U (r_i-1)y_i}{N\sum_Uy_i}$
where $r_i$ is the rank of $y_i$.

Note that it is not possible to use Taylor linearization for these measures because they rely on quantiles or, in the case of the Gini coefficient, a function of ranks.  Therefore, we instead follow the approach proposed by Deville (1999) based upon influence functions.


Let $U$ be a population of size $N$ and $M$ be a measure that allocates mass one to the set composed by one unit, that is $M(i)=M_i= 1$ if $i\in U$ and $M(i)=0$ if $i\notin U$.

Now, a population parameter $\theta$ can be expressed as a functional of $M$
$\theta=T(M)$.

Examples of such parameters are:

- Total: 
$Y=\sum_Uy_i=\sum_U y_iM_i=\int ydM=T(M)$

- Ratio of two totals:
$R=\frac{Y}{X}=\frac{\int y dM}{\int x dM}=T(M)$

- Cumulative distribution function:
$F(x)=\frac{\sum_U 1(y_i\leq x)}{N}=\frac{\int 1(y\leq x)dM}{\int{dM}}=T(M)$


To estimate these parameters from the sample, we replace the measure $M$ by the estimated measure $\hat{M}$ defined by: $\hat{M}(i)=\hat{M}_i= w_i$ if $i\in s$ and $\hat{M}(i)=0$ if $i\notin s$. 

The estimators of the population parameters can then be expressed as functional of the measure  $\hat{M}$. 

-  Total:
$\hat{Y}=T(\hat{M})=\int yd\hat{M}=\sum_s w_iy_i$

- Ratio of totals:
$\hat{R}=T(\hat{M})=\frac{\int y d\hat{M}}{\int x d\hat{M}}=\frac{\sum_s w_iy_i}{\sum_s w_ix_i}$

- Cumulative distribution function:
$\hat{F}(x)=T(\hat{M})=\frac{\int 1(y\leq x)d\hat{M}}{\int{d\hat{M}}}=\frac{\sum_s w_i 1(y_i\leq x)}{\sum_s w_i}$



## Influence Function Examples


- Total:
$$
\begin{aligned}
IT_k(M)&=lim_{t\rightarrow 0}\frac{T(M+t\delta_k)-T(M)}{t}\\
&=lim_{t\rightarrow 0}\frac{\int y.d(M+t\delta_k)-\int y.dM}{t}\\
&=lim_{t\rightarrow 0}\frac{\int yd(t\delta_k)}{t}=y_k	
\end{aligned}
$$


- Ratio of two totals:
$$
\begin{aligned}
IR_k(M)&=I\left(\frac{U}{V}\right)_k(M)=\frac{V(M)\times IU_k(M)-U(M)\times IV_k(M)}{V(M)^2}\\
&=\frac{X y_k-Y x_k}{X^2}=\frac{1}{X}(y_k-Rx_k)
\end{aligned}
$$

## Examples of Linearization Using the Influence Function

- At-risk-of-poverty threshold:
$$
arpt = 0.6\times m
$$
where $m$ is the median income.

$$
z_k= -\frac{0.6}{f(m)}\times\frac{1}{N}\times\left[I(y_k\leq m-0.5) \right]
$$


- At-risk-of-poverty rate:

\[
arpr=\frac{\sum_U I(y_i \leq t)}{\sum_U w_i}.100
\]
\[
z_k=\frac{1}{N}\left[I(y_k\leq t)-t\right]-\frac{0.6}{N}\times\frac{f(t)}{f(m)}\left[I(y_k\leq m)-0.5\right]
\]

where:

$N$ - population size; 

$t$ - at-risk-of-poverty threshold;

$y_k$ - income of person $k$;

$m$ - median income;

$f$ - income density function;


## Replication Designs

All major functions in the `convey` library have S3 methods  for the classes: `survey.design`, `svyrep.design` and `DBIdesign`. When the argument `design` is  a survey design with replicate weights created by the `survey` library, `convey` uses the method `svrepdesign`. 

Considering the remarks in [@W85],  p. 163, concerning the deficiency of the `Jackknife` method in estimating the variance of `quantiles`, we adopted the bootstrap method instead. 

The function `bootVar` from the `laeken` library [@R-laeken], also uses the bootstrap method to estimate variances. 

## Decomposition

Some inequality and multidimensional poverty measures can be decomposed. The decomposition methods in the `convey` library are limited to group decomposition.

For instance, the generalized entropy index can be decomposed into between and within group components. This sheds light on a very simple question: _of the overall inequality, how much can be explained by inequalities between groups and within groups?_ Since this measure is additively decomposable, one can get estimates of the coefficients, SEs and covariance between components. For a more practical approach, see [@lima2013].

The Alkire-Foster class of multidimensional poverty indices (not implemented in the `convey` library) can be decomposed by both dimensions and by groups, showing how much each group and each dimension contributes to poverty overall.  Multidimensional poverty techniques can sometimes help economic and policy analysts understand more precisely who is more affected by inequality and poverty, and how those disparities manifest.
