--- 
title: "Poverty and Inequality with Complex Survey Data"
author: "Guilherme Jacob, Anthony Damico, and Djalma Pessoa"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site

output:
  bookdown::tufte_html_book:
    toc: yes
    css: toc.css
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    latex_engine: xelatex
    citation_package: natbib
    keep_tex: yes
  bookdown::epub_book: default

    
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: guilhermejacob/context
description: "A book about the R convey package"
---

# Introduction

The R `convey` library estimates measures of poverty, inequality, and wellbeing.  There are two other R libraries covering this subject, [vardpoor](https://CRAN.R-project.org/package=vardpoor) and [laeken](https://CRAN.R-project.org/package=laeken), however, only `convey` integrates seamlessly with the [R survey package](https://CRAN.R-project.org/package=survey).

`convey` is free and open-source software that runs inside the [R environment for statistical computing](https://www.r-project.org/).  Anyone can review and propose changes to [the source code](https://github.com/DjalmaPessoa/convey) for this software.  Readers are welcome to [propose changes to this book](https://github.com/guilhermejacob/context/) as well.

## Installation {#install}

In order to work with the `convey` library, you will need to have R running on your machine.  If you have never used R before, you will need to [install that software](https://www.r-project.org/) before `convey` can be accessed.  Check out [FlowingData](http://flowingdata.com/2012/06/04/resources-for-getting-started-with-r/) for a concise list of resources for new R users.  Once you have R loaded on your machine, you can install..


* the latest released version from [CRAN](https://CRAN.R-project.org/package=convey) with

    ```R
    install.packages("convey")
    ````

* the latest development version from github with

    ```R
    devtools::install_github("djalmapessoa/convey")
    ```



## Complex surveys and statistical inference {#survey}

In this book, we demonstrate how to measure poverty and income concentration in a population based on microdata collected from a complex survey sample.  Most surveys administered by government agencies or larger research organizations utilize a sampling design that violates the assumption of simple random sampling (SRS), including:

1. Different units selection probabilities;
2. Clustering of units;
3. Stratification of clusters;
4. Reweighting to compensate for missing values and other adjustments.

Therefore, basic unweighted R commands such as `mean()` or `glm()` will not properly account for the weighting nor the measures of uncertainty (such as the confidence intervals) present in the dataset.  For some examples of publicly-available complex survey data sets, see [http://asdfree.com]().  

Unlike other software, the R `convey` package does not require that the user specify these parameters throughout the analysis.  So long as the [svydesign object](http://r-survey.r-forge.r-project.org/survey/html/svydesign.html) or [svrepdesign object](http://r-survey.r-forge.r-project.org/survey/html/svrepdesign.html) has been constructed properly at the outset of the analysis, the `convey` package will incorporate the survey design automatically and produce statistics and variances that take the complex sample into account.


## Usage Examples


In the following example, we've loaded the data set `eusilc` from the R libraries [vardpoor](https://CRAN.R-project.org/package=vardpoor) and [laeken](https://CRAN.R-project.org/package=laeken).

```{r results='hide', message=FALSE, warning=FALSE}
library(vardpoor)
data(eusilc)
```
Next, we create an object of class `survey.design` using the function `svydesign` of the library survey:

```{r results='hide', message=FALSE, warning=FALSE}
library(survey)
des_eusilc <- svydesign(ids = ~rb030, strata =~db040,  weights = ~rb050, data = eusilc)
```
Right after the creation of the design object `des_eusilc`, we should use the function `convey_prep` that adds an attribute to the survey design which saves information on the design object based upon the whole sample, needed to work with subset designs.

```{r}
library(convey)
des_eusilc <- convey_prep( des_eusilc )
```
To estimate the at-risk-of-poverty rate, we use the function `svyarpt`:

```{r comment=NA}
svyarpr(~eqIncome, design=des_eusilc)
```
To estimate the at-risk-of-poverty rate across domains defined by the variable `db040` we use:

```{r comment=NA}
svyby(~eqIncome, by = ~db040, design = des_eusilc, FUN = svyarpr, deff = FALSE)
```

Using the same data set, we estimate the quintile share ratio: 

```{r comment=NA}
# for the whole population
svyqsr(~eqIncome, design=des_eusilc, alpha= .20)

# for domains
svyby(~eqIncome, by = ~db040, design = des_eusilc,
  FUN = svyqsr, alpha= .20, deff = FALSE)

```

These functions can be used as S3 methods for the classes `survey.design` and `svyrep.design`.

Let's create a design object of class `svyrep.design` and run the function `convey_prep` on it:

```{r}
des_eusilc_rep <- as.svrepdesign(des_eusilc, type = "bootstrap")
des_eusilc_rep <- convey_prep(des_eusilc_rep) 
```

and then use the function `svyarpr`:

```{r comment=NA}
svyarpr(~eqIncome, design=des_eusilc_rep)

svyby(~eqIncome, by = ~db040, design = des_eusilc_rep, FUN = svyarpr, deff = FALSE)
```
The functions of the library convey are called in a similar way  to the functions in library survey.

It is also possible to deal with missing values by using the argument `na.rm`.

```{r comment=NA}
# survey.design using a variable with missings
svygini( ~ py010n , design = des_eusilc )
svygini( ~ py010n , design = des_eusilc , na.rm = TRUE )

# svyrep.design using a variable with missings
svygini( ~ py010n , design = des_eusilc_rep )
svygini( ~ py010n , design = des_eusilc_rep , na.rm = TRUE )
```



djalmapessoa_look, where do these references go on this page?  [@berger2003] and [@osier2009] and [@deville1999]



## Underlying Calculations


djalmapessoa_look, please describe the general purpose of linearization


In the `convey` library, there are some basic functions that produce the linearized variables needed to measure income concentration and poverty.  For example, looking at the income variable in some complex survey dataset, the `quantile` of that income variable can be linearized by the function `convey::svyiqalpha` and the sum total below any quantile of the variable is linearized by the function `convey::svyisq`.

From the linearized variables of these basic estimates, it is possible by using rules of composition, valid for influence functions, to derive the influence function of more complex estimates. By definition the influence function is a Gateaux derivative and the rules rules of composition valid for Gateaux derivatives also hold for Influence Functions.

The following property of Gateaux derivatives was often used in the library convey. Let $g$ be a differentiable function of $m$ variables. Suppose we want to compute the influence function of the estimator $g(T_1, T_2,\ldots, T_m)$, knowing the Influence function of the estimators $T_i, i=1,\ldots, m$. Then the following holds:

$$
I(g(T_1, T_2,\ldots, T_m)) = \sum_{i=1}^m \frac{\partial g}{\partial T_i}I(T_i)
$$

In the library convey this rule is implemented by the function `contrastinf` which uses the R function `deriv` to compute the formal partial derivatives $\frac{\partial g}{\partial T_i}$. 

For example, suppose we want to linearize the `Relative median poverty gap`(rmpg), defined as the difference between the at-risk-of-poverty threshold (`arpt`) and the median of incomes less than the `arpt` relative to the `arprt`: 

$$
rmpg= \frac{arpt-medpoor} {arpt}
$$

where `medpoor` is the median of incomes less than `arpt`.

Suppose we know how to linearize `arpt` and `medpoor`, then by applying the function `contrastinf` with 
$$
g(T_1,T_2)= \frac{(T_1 - T_2)}{T_1}
$$
we linearize the `rmpg`.



## The Variance Estimator


djalmapessoa_look please add references to this section


The variance of the estimator $T(\hat{M})$ can approximated by:

$$Var\left[T(\hat{M})\right]\cong var\left[\sum_s w_i z_i\right]$$

The  `linearized` variable $z$  is given by the derivative of the functional:

$$
z_k=lim_{t\rightarrow0}\frac{T(M+t\delta_k)-T(M)}{t}=IT_k(M)
$$

where, $\delta_k$ is the Dirac measure in $k$: $\delta_k(i)=1$ if and only if $i=k$.

This **derivative** is called  **Influence Function** and was introduced in the area of **Robust Statistics**.



## Influence Functions

Some measures of poverty and income concentration are defined by non-differentiable functions so that it is not possible to use Taylor linearization to estimate their variances. An alternative is to use **Influence functions** as described in [@deville1999] and [@osier2009]. The convey library implements this methodology to work with `survey.design` objects and also with `svyrep.design` objects.

Some examples of these measures are:

- At-risk-of-poverty threshold:
$arpt=.60q_{.50}$ where $q_{.50}$ is the income median;

- At-risk-of-poverty rate
$arpr=\frac{\sum_U 1(y_i \leq arpt)}{N}.100$

- Quintile share ratio
     
$qsr=\frac{\sum_U 1(y_i>q_{.80})}{\sum_U 1(y_i\leq q_{.20})}$

- Gini coefficient
$1+G=\frac{2\sum_U (r_i-1)y_i}{N\sum_Uy_i}$
where $r_i$ is the rank of $y_i$.

Note that it is not possible to use Taylor linearization for these measures because they depend on quantiles and the Gini is defined as a function of ranks. This could be done using the approach proposed by Deville (1999) based upon influence functions. 



Let $U$ be a population of size $N$ and $M$ be a measure that allocates mass one to the set composed by one unit, that is $M(i)=M_i= 1$ if $i\in U$ and $M(i)=0$ if $i\notin U$ 

Now, a population parameter $\theta$ can be expressed as a functional of $M$
$\theta=T(M)$

Examples of such parameters are:

- Total: 
$Y=\sum_Uy_i=\sum_U y_iM_i=\int ydM=T(M)$

- Ratio of two totals:
$R=\frac{Y}{X}=\frac{\int y dM}{\int x dM}=T(M)$

- Cumulative distribution function:
$F(x)=\frac{\sum_U 1(y_i\leq x)}{N}=\frac{\int 1(y\leq x)dM}{\int{dM}}=T(M)$


To estimate these parameters from the sample, we replace the measure $M$ by the estimated measure $\hat{M}$ defined by: $\hat{M}(i)=\hat{M}_i= w_i$ if $i\in s$ and $\hat{M}(i)=0$ if $i\notin s$. 

The estimators of the population parameters can then be expressed as functional of the measure  $\hat{M}$. 

-  Total:
$\hat{Y}=T(\hat{M})=\int yd\hat{M}=\sum_s w_iy_i$

- Ratio of totals:
$\hat{R}=T(\hat{M})=\frac{\int y d\hat{M}}{\int x d\hat{M}}=\frac{\sum_s w_iy_i}{\sum_s w_ix_i}$

- Cumulative distribution function:
$\hat{F}(x)=T(\hat{M})=\frac{\int 1(y\leq x)d\hat{M}}{\int{d\hat{M}}}=\frac{\sum_s w_i 1(y_i\leq x)}{\sum_s w_i}$



## Influence Function Examples


- Total:
$$
\begin{aligned}
IT_k(M)&=lim_{t\rightarrow 0}\frac{T(M+t\delta_k)-T(M)}{t}\\
&=lim_{t\rightarrow 0}\frac{\int y.d(M+t\delta_k)-\int y.dM}{t}\\
&=lim_{t\rightarrow 0}\frac{\int yd(t\delta_k)}{t}=y_k	
\end{aligned}
$$


- Ratio of two totals:
$$
\begin{aligned}
IR_k(M)&=I\left(\frac{U}{V}\right)_k(M)=\frac{V(M)\times IU_k(M)-U(M)\times IV_k(M)}{V(M)^2}\\
&=\frac{X y_k-Y x_k}{X^2}=\frac{1}{X}(y_k-Rx_k)
\end{aligned}
$$

## Examples of Linearization Using the Influence Function

- At-risk-of-poverty threshold:
$$
arpt = 0.6\times m
$$
where $m$ is the median income.

$$
z_k= -\frac{0.6}{f(m)}\times\frac{1}{N}\times\left[I(y_k\leq m-0.5) \right]
$$


- At-risk-of-poverty rate:

\[
 arpr=\frac{\sum_U I(y_i \leq t)}{\sum_U w_i}.100
\]
\[
z_k=\frac{1}{N}\left[I(y_k\leq t)-t\right]-\frac{0.6}{N}\times\frac{f(t)}{f(m)}\left[I(y_k\leq m)-0.5\right]
\]

where:

$N$ - population size; 

$t$ - at-risk-of-poverty threshold;

$y_k$ - income of person $k$;

$m$ - median income;

$f$ - income density function;


## Comparisons to Other Software

comparing se estimates for the domain "Tyrol"

```{r}
library(convey)
library(vardpoor)
data(eusilc)
names( eusilc ) <- tolower( names( eusilc ) )
library(survey)
```

function to estimate the se

```{r}
SE_lin2 <- function(t,des){
variance<-survey::svyrecvar(t/des$prob, des$cluster,des$strata, des$fpc,
postStrata = des$postStrata)
sqrt(variance)
}


des_eusilc <- svydesign(ids = ~rb030, strata =~db040,  weights = ~rb050, data = eusilc)

des_eusilc_Tyrol <- subset(des_eusilc, db040=="Tyrol")
```

total

convey

```{r}
incvar <- des_eusilc_Tyrol$variables$eqincome
lin.tot.conv <- incvar
SE_lin2 (lin.tot.conv, des_eusilc_Tyrol)
```

vardpoor

```{r}
ncom<- names(des_eusilc$prob)
ind <- names(des_eusilc_Tyrol$prob)
ID <- rep(1, nrow(des_eusilc))* (ncom %in% ind)
incvec <- des_eusilc$variables$eqincome
lin.tot.vardpoor <- incvec*ID
SE_lin2 (lin.tot.vardpoor, des_eusilc)
```

check using survey

```{r}
SE(svytotal(~eqincome, des_eusilc_Tyrol))
```

mean for domain or ratio:

convey

```{r}
num <- incvar
den <- rep(1, nrow(des_eusilc_Tyrol))
Y <- sum(num* weights(des_eusilc_Tyrol))
X <- sum(den* weights(des_eusilc_Tyrol))
R <- Y/X
```

linearized variable

```{r}
lin.ratio.conv <- (1/X)* (num-R*den )
SE_lin2 (lin.ratio.conv, des_eusilc_Tyrol)
```

vardpoor


```{r}
num <- incvec*ID
den <- rep(1, nrow(des_eusilc))*ID
Y <- sum(num* weights(des_eusilc))
X <- sum(den* weights(des_eusilc))
R <- Y/X
```
linearized variable

```{r}
lin.ratio.vardpoor <- (1/X)* (num-R*den )
SE_lin2 (lin.ratio.vardpoor, des_eusilc)
```

check by survey

```{r}
SE(svymean(~eqincome,des_eusilc_Tyrol ))
```

median

convey
```{r}
alpha <- .5
h <- h_fun(incvar, weights(des_eusilc_Tyrol))
q_alpha <- survey::svyquantile(~eqincome, des_eusilc_Tyrol, .5,
method = "constant")
q_alpha <- as.vector(q_alpha)
Fprime <- densfun( ~eqincome , des_eusilc_Tyrol, q_alpha,h, fun = "F")
N <- sum ( weights(des_eusilc_Tyrol))
lin.quant.conv <-  -(1/(N * Fprime)) * ((incvar <= q_alpha) - alpha)
```

se estimate
```{r}
SE_lin2 (lin.quant.conv, des_eusilc_Tyrol)
```

vardpoor

```{r}
lin.quant.vardpoor <-  -(1/(N * Fprime)) *ID* ((incvec <= q_alpha) - alpha)

SE_lin2 (lin.quant.vardpoor, des_eusilc)
```

Remarks: in all examples the results match. In these particular case, the linearized variable for vardpoor the ID always enter as a factor. There are formulas for the linearized variable in vardpoor where this is not the case. When this hapen, the se estimates are different!




## Replication Designs

djalmapessoa_look, please describe how the software works differently on svrepdesign objects -- as compared to svydesign objects

## Decomposition

Some inequality and multidimensional poverty measures can be decomposed. As of December 2016, the decomposition methods in `convey` are limited to group decomposition.

For instance, the generalized entropy index can be decomposed into between and within group components. This sheds light on a very simple question: of the overall inequality, how much can be explained by inequalities between groups and within groups? Since this measure is additive decomposable, one can get estimates of the coefficients, SEs and covariance between components. For a more practical approach, see [@lima2013].

The Alkire-Foster class of multidimensional poverty indices can be decomposed by dimension and groups. This shows how much each group (or dimension) contribute to the overall poverty.

This technique can help understand where and who is more affected by inequality and poverty, contributing to more specific policy and economic analysis.
